---
title: "Double-constrained Correspondence Analysis (dc-CA)"
params:
  colsite: "burlywood4"
  colspp: "cornflowerblue"
---


```{r, results='hide', message=FALSE}
#| code-fold: true

# Paths
library(here)

# Multivariate analysis
library(ade4)
library(adegraphics)

# dc-CA
source(here("functions/dpcaiv2-ade4.R"))

# Matrix algebra
library(expm)

# Plots
library(gridExtra)
source(here("functions/plot.R"))
```


*The contents of this page relies heavily on [@terbraak2018].*

## Introduction

dc-CA was developed as a natural extension of CCA and has been used to study the relationship between species traits and environmental variables.

In dc-CA, we have 3 matrices:

+ A data matrix $Y$ ($r \times c$)
+ A matrix of predictor variables $E$ ($r \times l$)
+ A matrix of predictor variables $T$ ($c \times k$)


```{r, echo = FALSE, message = FALSE}
dat <- readRDS(here("data/Barbaro2012.rds"))
Y <- dat$comm
E <- dat$envir
T_ <- dat$traits

r <- dim(Y)[1]
c <- dim(Y)[2]
l <- dim(E)[2]
k <- dim(T_)[2]

plotmat(r = r, c = c, 
        E = TRUE, T_ = TRUE,
        l = l, k = k)
```

The aim of dc-CA is to find a linear combination of the predictor variables in $E$ and $T$ (environmental variables and traits) that maximizes the correlation.

Below are these matrices for our data:

$Y =$
```{r, echo = FALSE, message = FALSE}
knitr::kable(Y)
```


$E =$
```{r, echo = FALSE, message = FALSE}
knitr::kable(E)
```

$T =$
```{r, echo = FALSE, message = FALSE}
knitr::kable(T_)
```

```{r}
(r <- dim(Y)[1])
(c <- dim(Y)[2])
(l <- dim(E)[2])
(k <- dim(T_)[2])
```

dc-CA must not have to many traits compared to species: that is a disadvantage compared to RLQ, but on the other hand dc-CA allows to see relationships that RLQ would miss [@terbraak2018].

There are several ways to perform dc-CA [@terbraak2018], notably:

-   singular value decomposition (the method used here)
-   an iterative method *à la* reciprocal averaging
-   canonical correlation analysis between $T$ et $E$, weighted by $Y$

## Computation

::: {.callout-tip title="*TL;DR*"}
We perform the SVD of $D = [E_{center}' D(y_{i \cdot}) E_{center}]^{-1/2} E_{center}' Y T_{center} [T_{center}' D(y_{\cdot j}) T_{center}]^{-1/2}$

$$
D = P \Delta Q'
$$
This allows us to find the eigenvectors $P$ (rows or environment eigenvectors) and $Q$ (columns or traits eigenvectors).

The eigenvalues of the dc-CA are the squared eigenvalues of the SVD: $\Lambda = \Delta^2$.

There are $\min(k, l)$ non-null eigenvalues???

**Regression coefficients**

Then, the regression coefficients for explanatory variables can be found with:

+ $B = [E_{center}'D(y_{i\cdot})E_{center}]^{-1/2}P\Delta^\alpha$ (row explanatory variables = environmental variables)
+ $C = [T_{center}' D(y_{\cdot j})T_{center}]^{-1/2}Q\Delta^\alpha$ (columns explanatory variables = species traits)

where $\alpha$ is a scaling factor equal to 1, 0 or 1/2. The different scalings are detailed in the next part.

**Individuals coordinates**

The individuals coordinates (species or sites) can be computed in two ways:

Linear combinations (LC scores) are computed from those coefficients : 

+ $Z_i = E_{center}B_i$ for the rows (sites) 
+ $\hat{Z}_i = T_{center}C_i$ for columns (species)

Weighted averages (WA scores) are computed from the scores of the **other** individuals:

+ $\hat{U}_i = D(y_{\cdot i})^{-1} Y Z_i$ for row (sites) scores 
+ $U_i = D(y_{j \cdot})^{-1} Y' \hat{Z}_i$ for column (species) scores

**Correlations**

Finally, we can compute correlations scores for different scalings for the explanatory variables:

+ $BS_B^i = \left[ E_{center}'D(y_{i\cdot})E_{center} \right]^{1/2} P \Delta^{\alpha}$ for the row (environmental variables)
+ $BS_C^i = \left[ T_{center}'D(y_{i\cdot})T_{center} \right]^{1/2} Q \Delta^{1-\alpha}$ for the columns (species traits)
:::


First, we need to center the traits and environment matrices (resp. $T_{cent}$ and $E_{cent}$). To do that, we have to compute "inflated" versions of these matrices matching the occurrence counts in $Y$.

$$
E_{stand} = E - \bar{E}_{infl} =\left[e_i \sum_i y_{i\cdot}e_i/y_{\cdot\cdot} \right]
$$

$$
T_{stand} = T - \bar{T}_{infl} = \left[t_j \sum_j y_{\cdot j}t_j/y_{\cdot\cdot} \right]
$$

With our data:

```{r}
# Center E -----
yi_ <- rowSums(Y)
Ecenter <- matrix(nrow = nrow(E), ncol = ncol(E))

for(i in 1:ncol(Ecenter)) {
  Ecenter[, i] <- E[, i] - sum(E[, i]*yi_)/sum(Y)
}
# This is the same as computing a mean on inflated data matrix Einfl and centering E with these means

# Center T -----
y_j <- colSums(Y)
Tcenter <- matrix(nrow = nrow(T_), ncol = ncol(T_))
rownames(Tcenter) <- rownames(T_)
colnames(Tcenter) <- colnames(T_)

for(j in 1:ncol(Tcenter)) {
  Tcenter[, j] <- T_[, j] - (sum(T_[, j]*y_j)/sum(Y))
}
```

```{r}
# Check centering -----
M1 <- matrix(rep(1, nrow(Y)), nrow = 1)
all((M1 %*% diag(rowSums(Y)) %*% Ecenter) < 10e-10)

M1 <- matrix(rep(1, ncol(Y)), nrow = 1)
all((M1 %*% diag(colSums(Y)) %*% Tcenter) < 10e-10)
```

Maximizing the fourth-corner correlation means finding $\hat{u}$ and $u$ that maximize $\hat{u}'Yu$ (where $\hat{u}$ are the sites (rows) scores and $u$ are the species (columns) scores).

$\hat{u}$ and $u$ are linear combinations of traits and environmental variables: $\hat{u} = E_{center}b$ and $u = T_{center}c$. 

So in the end, we need to maximize $\hat{u}'Yu$ with respect to the coefficients vectors $b$ and $c$:

$$
\max_{b, c}(\hat{u}'Yu) = \max_{b, c}\left(\left[E_{center}b\right]'Y  T_{center}c \right) 
$$

::: {.callout-note}
These equations are written for the first axis, but we can also write them in matrix form:

$$
\max_{B, C}(\hat{U}'YU) = \max_{B, C}\left(\left[E_{center}B\right]'Y T_{center}C \right) 
$$
::: 

In addition, we introduce the following constraint on the norm of the columns vectors of $\hat{U}$ and $U$: $\hat{u}' D(y_{i\cdot}) \hat{u} = 1$ and $u' D(y_{\cdot j})u = 1$. In fact, tjese constraints will be relaxed later depending on the scaling (see below).

To find the coefficients $B$ and $C$ defined above, we need to diagonalize the following matrices $M$ and $M_2$:

+ To find $B$, we must diagonalize $M$:
$$
M = \left[E_{center}' D(y_{i\cdot}) E_{center} \right]^{-1} E_{center}'YT_{center} \left[T_{center}' D(y_{\cdot j}) T_{center} \right]^{-1} T_{center}' Y' E_{center}
$$
Where matrices $D(y_{i \cdot})$ and $D(y_{\cdot j})$ are the diagonal matrices with the column and row sums (respectively). They are analogous to matrices $D(p_{i \cdot})$ and $D(p_{\cdot j})$ defined for CA and CCA (but we use $Y$ instead of $P$ to define the margins).


We can view $M$ as:
$$
M = \hat{E}_{center} \hat{T}_{center} = \beta T_{center} \gamma E_{center}
$$
With $\beta = \left[E_{center}' D(y_{i\cdot}) E_{center} \right]^{-1} E_{center}'Y$ and $\gamma = \left[T_{center}' D(y_{\cdot j}) T_{center} \right]^{-1} T_{center}' Y'$. Here, we predict the environment with traits so that $\hat{E}_{center} = \beta T_{center}$ and the traits with the environment so that $\hat{T}_{center} = \gamma E_{center}$.
So we maximize the correlation between $\hat{E}_{center}$ and $\hat{T}_{center}$.

+ Similarly, to find $C$, we must diagonalize $M_2$:

$$
M_2 = \left[T_{center}' D(y_{\cdot j}) T_{center} \right]^{-1} T_{center}' Y' E_{center} \left[E_{center}' D(y_{i\cdot}) E_{center} \right]^{-1} E_{center}'YT_{center}
$$
Here, we can view $M_2$ as:
$$
M_2 = \hat{T}_{center} \hat{E}_{center} = \gamma E_{center} \beta T_{center}
$$

The eigenvectors matrices of these diagonalizations give us $B$ and $C$:

$$
M = B \Lambda_b B^{-1} ~~ \text{and} ~~ M_2 = C \Lambda_c C^{-1}
$$

We can either diagonalize $M$ and $M_2$ of perform a single SVD of a matrix $D$ defined below and get a similar result. Both methods are presented below.

### Diagonalizations

We diagonalize $M$ and $M_2$:

```{r}
# Define weights
Dyi_ <- diag(rowSums(Y))
Dy_j <- diag(colSums(Y))

# Compute M
M <- solve(t(Ecenter) %*% Dyi_ %*% Ecenter) %*% t(Ecenter) %*% Y %*% Tcenter %*% solve(t(Tcenter) %*% Dy_j %*% Tcenter) %*% t(Tcenter) %*% t(Y) %*% Ecenter

# Compute M2
M2 <- solve(t(Tcenter) %*% Dy_j %*% Tcenter) %*% t(Tcenter) %*% t(Y) %*% Ecenter %*% solve(t(Ecenter) %*% Dyi_ %*% Ecenter) %*% t(Ecenter) %*% Y %*% Tcenter 
```


```{r}
# Diagonalize M
eigB <- eigen(M)
lambdaB <- eigB$values
lambdaB # All non-null eigenvectors
vB <- eigB$vectors

# Diagonalize M2
eigC <- eigen(M2)
lambdaC <- eigC$values
lambdaC # six non-null eigenvalues
vC <- eigC$vectors

all(lambdaB - lambdaC[1:l] < 10e-10)
```


### SVD

Alternatively, we can compute the SVD of a matrix $D$ computed from $M$ and get a similar result. $D$ is defined as:

$$
D = M^{1/2} = [E_{center}' D(y_{i \cdot}) E_{center}]^{-1/2} E_{center}' Y T_{center} [T_{center}' D(y_{\cdot j}) T_{center}]^{-1/2}
$$ 

With our dataset:


```{r}
D <- solve(sqrtm(t(Ecenter) %*% Dyi_ %*% Ecenter)) %*%
  t(Ecenter) %*% Y %*% Tcenter %*%
  solve(sqrtm(t(Tcenter) %*% Dy_j %*% Tcenter))
```

We perform the SVD of $D$: $D = P \Lambda_{SVD} Q'$.

```{r}
sv <- svd(D)

delta <- sv$d
Delta <- diag(delta)

P_svd <- sv$u
Q_svd <- sv$v

dim(P_svd)
dim(Q_svd)
```


The eigenvalues of the SVD are the square roots of the final eigenvalues obtained by diagonalization: $\Lambda = \Delta^2$

```{r}
lambda <- delta^2
Lambda <- diag(lambda)
```


## Scalings

::: {.callout-tip title="*TL;DR*"}
There are two types of coordinates: linear combination scores (LC scores) and weighted averages scores (WA scores).

**The general formulas are:**

+ the regression coefficients for the rows (environmental variables) are $B = [E_{center}'D(y_{i\cdot})E_{center}]^{-1/2}P\Delta^\alpha$
+ the regression coefficients for the columns (species traits) are $C = [T_{center}' D(y_{\cdot j})T_{center}]^{-1/2}Q\Delta^{1 - \alpha}$
+ the LC scores for rows (sites) are $Z_i = E_{center}B_i$
+ the LC scores for columns (species) are $\hat{Z}_i = T_{center}C_i$
+ the WA scores for rows (sites) are $\hat{U}_i = D(y_{\cdot i})^{-1} Y Z_i$
+ the WA scores for columns (species) are $U_i = D(y_{j \cdot})^{-1} Y' \hat{Z}_i$
+ the correlations with axes for the rows (environmental variables) are $BS_B^i = \left[ E_{center}'D(y_{i\cdot})E_{center} \right]^{1/2} P \Delta^{\alpha}$
+ the correlations with axes for the columns (species traits) are $BS_C^i = \left[ T_{center}'D(y_{i\cdot})T_{center} \right]^{1/2} Q \Delta^{1-\alpha}$


***Note: the formula for the regression coefficients $C_i$ are in disagreement with the article of [@terbraak2018]. Here, we find that the scaling should include $\Delta^{1-\alpha}$ in the formula (17) instead of $\Delta^{\alpha-1}$. However, when we try to prove transition formula (6), we find that it should be $\Delta^{\alpha}$.***


In these formulas, note that the WA scores for one dimension are computed from the predicted scores of the other dimension.

**Scaling type 1 ($\alpha = 1$)**

+ the regression coefficients for the rows (environmental variables) are $B_1 = [E_{center}'D(y_{i\cdot})E_{center}]^{-1/2}P\Delta$
+ the regression coefficients for the columns (species traits) are $C_1 = [T_{center}' D(y_{\cdot j})T_{center}]^{-1/2}Q$
+ the LC scores for rows (sites) are $Z_1 = E_{center}B_1$
+ the LC scores for columns (species) are $\hat{Z}_1 = T_{center}C_1$
+ the WA scores for rows (sites) are $\hat{U}_1 = D(y_{\cdot i})^{-1} Y Z_1$
+ the WA scores for columns (species) are $U_1 = D(y_{j \cdot})^{-1} Y' \hat{Z}_1$
+ the correlations with axes for the rows (environmental variables) are $BS_B^1 = \left[ E_{center}'D(y_{i\cdot})E_{center} \right]^{1/2} P \Delta$
+ the correlations with axes for the columns (species traits) are $BS_C^1 = \left[ T_{center}'D(y_{i\cdot})T_{center} \right]^{1/2} Q \Delta$


**Scaling type 2 ($\alpha = 0$)**

+ the regression coefficients for the rows (environmental variables) are $B_2 = [E_{center}'D(y_{i\cdot})E_{center}]^{-1/2}P$
+ the regression coefficients for the columns (species traits) are $C_2 = [T_{center}' D(y_{\cdot j})T_{center}]^{-1/2}Q \Delta$
+ the LC scores for rows (sites) are $Z_2 = E_{center}B_2$
+ the LC scores for columns (species) are $\hat{Z}_2 = T_{center}C_2$
+ the WA scores for rows (sites) are $\hat{U}_2 = D(y_{\cdot i})^{-1} Y Z_2$
+ the WA scores for columns (species) are $U_2 = D(y_{j \cdot})^{-1} Y' \hat{Z}_2$
+ the correlations with axes for the rows (environmental variables) are $BS_B^2 = \left[ E_{center}'D(y_{i\cdot})E_{center} \right]^{1/2} P$
+ the correlations with axes for the columns (species traits) are $BS_C^2 = \left[ T_{center}'D(y_{i\cdot})T_{center} \right]^{1/2} Q$

**Scaling type 3 ($\alpha = 1/2$)**

+ the regression coefficients for the rows (environmental variables) are $B_3 = [E_{center}'D(y_{i\cdot})E_{center}]^{-1/2}P\Delta^{1/2}$
+ the regression coefficients for the columns (species traits) are $C_3 = [T_{center}' D(y_{\cdot j})T_{center}]^{-1/2}Q\Delta^{1/2}$
+ the LC scores for rows (sites) are $Z_3 = E_{center}B_3$
+ the LC scores for columns (species) are $\hat{Z}_3 = T_{center}C_3$
+ the WA scores for rows (sites) are $\hat{U}_3 = D(y_{\cdot i})^{-1} Y Z_3$
+ the WA scores for columns (species) are $U_3 = D(y_{j \cdot})^{-1} Y' \hat{Z}_3$
+ the correlations with axes for the rows (environmental variables) are $BS_B^3 = \left[ E_{center}'D(y_{i\cdot})E_{center} \right]^{1/2} P \Delta^{1/2}$
+ the correlations with axes for the columns (species traits) are $BS_C^3 = \left[ T_{center}'D(y_{i\cdot})T_{center} \right]^{1/2} Q \Delta^{1/2}$


$\alpha$ changes the interpretation of the correlations vectors: 

+ intra-set correlations: $BS_B^2$ ($\alpha = 0$) approximates the correlations between the environmental variables (rows variables) and $BS_C^1$ ($\alpha = 1$) approximates the correlations between the traits (columns variables)
+ inter-set correlations (fourth-corner): $BS_B^1$ ($\alpha = 1$) approximates the correlation between environmental variables (row variables) and species (columns) and $BS_C^2$ ($\alpha = 0$) approximates the correlation between traits (columns variables) and sites (rows)
+ $BS_B^3$ and $BS_C^3$ ($\alpha = 1/2$) is the geometric mean of scalings 1 and 2.

When plotting the correlation circle, to look at the correlations between variables of the same set (traits or environmental variables), we should use $BS_B^2$ and $BS_C^1$. There are the scores returned by `ade4`.
:::


```{r}
ca <- dudi.coa(Y, 
               nf = c-1, 
               scannf = FALSE)

dcca <- dpcaiv2(dudi = ca, 
                dfR = E,
                dfQ = T_,
                scannf = FALSE, 
                nf = min(k, l))
```

### Pre-computations

Coefficients etc computed with our method are equal to those computed by `ade4`, but we need to multiply them by a scaling factor $\sqrt{y_{\cdot \cdot}}$.

```{r}
(scaling <- sqrt(sum(Y)))
```


We will need this function to normalize correlations vectors to compare results to those of `ade4`.

```{r}
#' Normalize row or columns vectors of a matrix
#'
#' @param M the matrix to normalize
#' @param margin the margin (1 = rows, 2 = columns)
#'
#' @return The normalized matrix M
normalize <- function(M, margin) {
  
  m_norm <- apply(M,
                  margin, 
                  function(x) sqrt(sum(x^2)))
  M_norm <- sweep(M, margin, m_norm, "/")
  return(M_norm)
}
```


### Scaling type 1

This type of scaling preserves the distances between rows. 

+ the rows (sites) scores can be $\hat{U}_1$ (WA scores) or $\hat{Z}_1$ (LC scores)  
+ the columns (species) scores can be $U_1$ (WA scores) or $Z_1$ (LC scores) 
+ rows variables (environmental variables) correlations are $BS_{B1}$.
+ columns variables (species traits) correlations are $BS_{C1}$


With the scaling type 1, $BS_{B1}$ represents the correlation between environmental variables and species and $BS_{C1}$ represents the correlation between species traits and species.

```{r}
# Coefficients
B_1 <- solve(sqrtm(t(Ecenter) %*% Dyi_ %*% Ecenter)) %*%  P_svd %*% Delta
C_1 <- solve(sqrtm(t(Tcenter) %*% Dy_j %*% Tcenter)) %*% Q_svd

# LC scores
Zhat1 <- Ecenter %*% B_1 # rows
Z1 <- Tcenter %*% C_1 # columns

# WA scores
Uhat1 <- solve(Dyi_) %*% Y %*% Z1 # rows
U1 <- solve(Dy_j) %*% t(Y) %*% Zhat1  # columns

# Variables scores
BS_B1 <- sqrtm(t(Ecenter) %*% Dyi_ %*% Ecenter) %*% P_svd %*% Delta
BS_C1 <- sqrtm(t(Tcenter) %*% Dy_j %*% Tcenter) %*% Q_svd

# Normalize
BS_B1norm <- normalize(BS_B1, 1)
BS_C1norm <- normalize(BS_C1, 1)
```

```{r, include = FALSE}
# Compare diagonalization and SVD
B_1/vB # The eigenvectors are almost the same
apply(B_1, 2, function(x) sqrt(sum(x^2))) # column vectors of B have different norms
apply(vB, 2, function(x) sqrt(sum(x^2))) # column vectors of vB are of norm 1

C_1/vC[, 1:l] # The eigenvectors are almost the same
apply(C_1, 2, function(x) sqrt(sum(x^2))) # column vectors of C have different norms
apply(vC, 2, function(x) sqrt(sum(x^2))) # column vectors of vC are of norm 1
```

```{r, include = FALSE}
Z1*scaling/dcca$c1
Zhat1*scaling/dcca$li

Uhat1*scaling/dcca$lsR
BS_C1norm/dcca$corQ # not exactly equal to one -> numeric issue?
```


```{r, include = FALSE}
s.label(dcca$c1, # Species constrained at variance 1
        plabels.optim = TRUE,
        plabels.col  = params$colspp,
        ppoints.col = params$colspp)
s.label(dcca$li, # Sites position by averaging
        plabels.optim = TRUE,
        plabels.col  = params$colsite,
        ppoints.col = params$colsite,
        add = TRUE)
s.corcircle(dcca$corQ)
```


```{r, fig.height=12, fig.width=6, warning=FALSE}
#| code-fold: true

# WA scores
gwa <- multiplot(indiv_row = Uhat1*scaling, indiv_col = U1*scaling, 
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 var_row = BS_B1norm, var_row_lab = colnames(E),
                 var_col = BS_C1norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("WA scores for sites")

# LC scores
glc <- multiplot(indiv_row = Zhat1*scaling, indiv_col = Z1*scaling, 
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 var_row = BS_B1norm, var_row_lab = colnames(E),
                 var_col = BS_C1norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("LC scores for sites")

grid.arrange(grobs = list(gwa, glc),
             nrow = 2)
```


On this plot:

+ row and columns variables ($BS_B^1$ - $BS_C^1$): angles between arrows of rows and environmental variables represent their fourth-corner correlation. For instance, the correlation between `mass` and `perc_forests` is large (arrows size + direction).
+ species traits ($BS_C^1$): arrows indicate intra-set correlations.
+ species and environmental variables ($U$ - $BS_B^1$): species points on the right are located in sites with a high `location` variable.
+ sites and species traits ($\hat{U}$ - $BS_C^1$): we can understand sites using their community weighted mean (eg site 26 has birds with high mass).
+ species and species traits ($U$ - $BS_C^1$): for instance, species 12 has large eggs.

**There is no interpretation for sites - environmental variables ($\hat{U}$ - $BS_B^1$).**

### Scaling type 2

This type of scaling preserves the distances between columns.

+ the rows (sites) scores can be $\hat{U}_2$ (WA scores) or $\hat{Z}_2$ (LC scores)  
+ the columns (species) scores can be $U_2$ (WA scores) or $Z_2$ (LC scores) 
+ rows variables (environmental variables) correlations are $BS_{B2}$.
+ columns variables (species traits) correlations are $BS_{C2}$


With the scaling type 2, $BS_{B2}$ represents the correlation between environmental variables and sites and $BS_{C2}$ represents the correlation between species traits and sites.

```{r}
# Regression coefficients
B_2 <- solve(sqrtm(t(Ecenter) %*% Dyi_ %*% Ecenter)) %*%  P_svd
C_2 <- solve(sqrtm(t(Tcenter) %*% Dy_j %*% Tcenter)) %*% Q_svd %*% Delta

# LC scores
Zhat2 <- Ecenter %*% B_2 # rows
Z2 <- Tcenter %*% C_2 # columns

# WA scores
Uhat2 <- solve(Dyi_) %*% Y %*% Z2 # rows
U2 <- solve(Dy_j) %*% t(Y) %*% Zhat2 # columns

# Variables scores
BS_B2 <- sqrtm(t(Ecenter) %*% Dyi_ %*% Ecenter) %*% P_svd
BS_C2 <- sqrtm(t(Tcenter) %*% Dy_j %*% Tcenter) %*% Q_svd %*% Delta

# Normalize
BS_B2norm <- normalize(BS_B2, 1)
BS_C2norm <- normalize(BS_C2, 1)
```

```{r, include = FALSE}
Z2*scaling/dcca$co
Zhat2*scaling/dcca$l1

U2*scaling/dcca$lsQ
BS_B2norm/dcca$corR
```

```{r, fig.height=12, fig.width=6, warning=FALSE}
#| code-fold: true

# WA scores
gwa <- multiplot(indiv_row = Uhat2*scaling, indiv_col = U2*scaling, 
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 var_row = BS_B2norm, var_row_lab = colnames(E),
                 var_col = BS_C2norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("WA scores for sites")

# LC scores
glc <- multiplot(indiv_row = Zhat2*scaling, indiv_col = Z2*scaling, 
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 var_row = BS_B2norm, var_row_lab = colnames(E),
                 var_col = BS_C2norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("LC scores for sites")

grid.arrange(grobs = list(gwa, glc),
             nrow = 2)
```

On this plot:

+ row and columns variables ($BS_B^2$ - $BS_C^2$): angles between arrows of rows and environmental variables represent their fourth-corner correlation. For instance, the correlation between `mass` and `patch_forests` is large (arrows size + direction).
+ environmental variables ($BS_B^2$): arrows indicate intra-set correlations.
+ species and environmental variables ($U$ - $BS_B^1$)
+ sites and species traits ($\hat{U}$ - $BS_C^1$)
+ sites and environmental variables ($U$ - $BS_C^1$)

**There is no interpretation for species - species traits ($U$ - $BS_C^2$).**

### Scaling type 3

This type of scaling is an intermediate between scalings 1 and 2.

+ the rows (sites) scores can be $\hat{U}_3$ (WA scores) or $\hat{Z}_3$ (LC scores)  
+ the columns (species) scores can be $U_3$ (WA scores) or $Z_3$ (LC scores) 
+ rows variables (environmental variables) correlations are $BS_{B3}$.
+ columns variables (species traits) correlations are $BS_{C3}$

With the scaling type 3, $BS_{B3}$ and $BS_{C3}$ represent the geometric mean of their correlation with species and sites.

```{r}
# Variables coefficients
B_3 <- solve(sqrtm(t(Ecenter) %*% Dyi_ %*% Ecenter)) %*%  P_svd %*% Delta^(1/2)
C_3 <- solve(sqrtm(t(Tcenter) %*% Dy_j %*% Tcenter)) %*% Q_svd %*% Delta^(1/2)

# LC scores
Zhat3 <- Ecenter %*% B_3 # rows
Z3 <- Tcenter %*% C_3 # columns

# WA scores
Uhat3 <- solve(Dyi_) %*% Y %*% Tcenter %*% C_3 # rows
U3 <- solve(Dy_j) %*% t(Y) %*% Zhat3 # columns

# Variables scores
BS_B3 <- sqrtm(t(Ecenter) %*% Dyi_ %*% Ecenter) %*% P_svd %*% Delta^(1/2)
BS_C3 <- sqrtm(t(Tcenter) %*% Dy_j %*% Tcenter) %*% Q_svd %*% Delta^(1/2)

# Normalize
BS_B3norm <- normalize(BS_B3, 1)
BS_C3norm <- normalize(BS_C3, 1)
```

```{r, fig.height=12, fig.width=6, warning=FALSE}
#| code-fold: true

# WA scores
gwa <- multiplot(indiv_row = Uhat3*scaling, indiv_col = U3*scaling, 
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 var_row = BS_B3norm, var_row_lab = colnames(E),
                 var_col = BS_C3norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("WA scores for sites")

# LC scores
glc <- multiplot(indiv_row = Zhat3*scaling, indiv_col = Z3*scaling, 
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 var_row = BS_B3norm, var_row_lab = colnames(E),
                 var_col = BS_C3norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("LC scores for sites")

grid.arrange(grobs = list(gwa, glc),
             nrow = 2)
```


## More tests

Below, we test the variances of various scores.

```{r}
# Since ter Braak and ade4 disagree, check the variance of co and Z2 or Z2 (it should be the eigenvalues of each axis)

#' Get the variance of vector x
#'
#' @param x The vector
#' @param w The weights
#'
#' @return The variance
varvec <- function(x, w) {
  sum(w*(x-mean(x))^2)
}

res_ade4 <- apply(as.matrix(dcca$co), 
                  2, 
                  function(x) varvec(x, dcca$cw))
res_ade4/dcca$eig
  
res_tB <- apply(Z2*scaling, 
                2, 
                function(x) varvec(x, dcca$cw))
res_tB/dcca$eig 

res_agree <- apply(Z2*scaling, 
                   2, 
                   function(x) varvec(x, dcca$cw))
res_agree/dcca$eig
```

+ LC scores for rows:
```{r}
# l1 variances should be 1
apply(as.matrix(dcca$l1), 
      2, 
      function(x) varvec(x, dcca$lw))

# li variances should be eigenvalues
apply(as.matrix(dcca$li), 
      2, 
      function(x) varvec(x, dcca$lw))/dcca$eig
```
+ LC scores for columns
```{r}
# c1 variances should be 1
apply(as.matrix(dcca$c1), 
      2, 
      function(x) varvec(x, dcca$cw))

# co variances should be eigenvalues
apply(as.matrix(dcca$co), 
      2, 
      function(x) varvec(x, dcca$cw))/dcca$eig 
```
```{r}
# Z2 variances should be the eigenvalues
apply(Z2*scaling, 
      2, 
      function(x) varvec(x, dcca$cw))/dcca$eig
```
+ WA scores:
```{r}
# Hypothesis: these scires shound be the eigenvalues
apply(as.matrix(dcca$lsR), 
      2, 
      function(x) varvec(x, dcca$lw))/dcca$eig

# co variances should be eigenvalues
apply(as.matrix(dcca$lsQ), 
      2, 
      function(x) varvec(x, dcca$cw))/dcca$eig
```

## Interpretation

This method finds the linear correlation of row explanatory variables (environmental variables) and the linear correlation of columns explanatory variables (species traits) that maximizes the fourth-corner correlation, i.e. the correlation between these linear combinations of row and columns-variables.

There are other related methods, that have been better described and also more used in ecology: RLQ, community weighted means RDA (CMW-RDA). 

Contrary to RLQ, dc-CA takes into account the correlation between the row and column variables. Thus, while RLQ can analyze any number of row and column variables, it is not the case with dc-CA the number of row and column variables must not be large compared to the number of rows/columns in the tables. Also, CCA maximizes correlation and RLQ maximizes covariance [@terbraak2018].

The eigenvalues of dc-CA are the squares of the fourth-corner correlations.