---
title: "Double constrained Correspondence Analysis (dc-CA)"
params:
  colsite: "burlywood4"
  colspp: "cornflowerblue"
---


```{r, results='hide', message=FALSE}
#| code-fold: true

# Paths
library(here)

# Multivariate analysis
library(ade4)
library(adegraphics)

# Matrix algebra
library(expm)

# Plots
library(CAnetwork)
library(patchwork)
library(ggplot2)
```


```{r}
#| code-fold: true

# Define bound for comparison to zero
zero <- 10e-10
```


```{r}
#| code-fold: true


#' Normalize row or columns vectors of a matrix
#'
#' @param M the matrix to normalize
#' @param margin the margin (1 = rows, 2 = columns)
#'
#' @return The normalized matrix M
normalize <- function(M, margin) {
  
  m_norm <- apply(M,
                  margin, 
                  function(x) sqrt(sum(x^2)))
  M_norm <- sweep(M, margin, m_norm, "/")
  return(M_norm)
}
```

*The contents of this page relies heavily on @terbraak2018.*


## Introduction

Double constrained correspondence analysis (dc-CA) was developed as a natural extension of CCA and has been used to study the relationship between species traits and environmental variables.

In dc-CA, we have 3 matrices:

+ A data matrix $Y$ ($r \times c$)
+ A matrix of predictor variables $E$ ($r \times l$)
+ A matrix of predictor variables $T$ ($c \times k$)


```{r, message = FALSE}
#| code-fold: true
dat <- readRDS(here("data/Barbaro2012.rds"))
Y <- dat$comm
E <- dat$envir
T_ <- dat$traits

r <- dim(Y)[1]
c <- dim(Y)[2]
l <- dim(E)[2]
k <- dim(T_)[2]

plotmat(r = r, c = c, 
        E = TRUE, T_ = TRUE,
        l = l, k = k)
```

The aim of dc-CA is to find a linear combination of the predictor variables in $E$ and $T$ (environmental variables and traits) that maximizes the correlation.

Below are the first lines of these matrices for our data:

$Y =$
```{r}
#| code-fold: true
knitr::kable(head(Y))
```


$E =$
```{r}
#| code-fold: true
knitr::kable(head(E))
```

$T =$
```{r}
#| code-fold: true
knitr::kable(head(T_))
```

```{r}
(r <- dim(Y)[1])
(c <- dim(Y)[2])
(l <- dim(E)[2])
(k <- dim(T_)[2])
```

dc-CA must not have to many traits compared to species: that is a disadvantage compared to RLQ, but on the other hand dc-CA allows to see relationships that RLQ would miss [@terbraak2018].

There are several ways to perform dc-CA [@terbraak2018], notably:

-   singular value decomposition (the method used here)
-   an iterative method *à la* reciprocal averaging
-   canonical correlation analysis between $T$ et $E$, weighted by $P$

## Computation

::: {.callout-tip title="*TL;DR*"}

We define matrix $D$ as: 

$$D = [\tilde{E}^\top D_r \tilde{E}]^{-1/2} \tilde{E}^\top P_0 \tilde{T} [\tilde{T}^\top D_c \tilde{T}]^{-1/2}$$

*Note: contrary to @terbraak2018, in this document we use the matrix $P_0$ instead of $Y$.*

We perform the SVD of $D$:

$$
D = B_0 \Delta C_0^\top
$$

This allows us to find the eigenvectors $B_0$ (regression coefficients multiplying rows/environment variables) and $C_0$ (regression coefficients multiplying columns/species traits).

The eigenvalues of the dc-CA are the squared eigenvalues of the SVD: $\Lambda = \Delta^2$.

There are $\min(k, l)$ non-null eigenvalues???

**Transformation of $B_0$ and $C_0$**

Then, we transform $B_0$ and $C_0$ (scaling):

+ $B = [\tilde{E}^\top D_r \tilde{E}]^{-1/2} B_0$ (coefficients for rows/environmental variables)
+ $C = [\tilde{T}^\top D_c \tilde{T}]^{-1/2} C_0$ (coefficients for columns/species traits)

**Individuals coordinates**

The individuals coordinates (species or sites) can be computed in two ways:

Linear combinations (LC scores) are computed from those coefficients : 

+ $U = \tilde{E} B$ for the rows (sites) 
+ $V = \tilde{T} C$ for columns (species)

Weighted averages (WA scores) are computed from the scores of the **other** individuals:

+ $U^\star = {D_r}^{-1} P_0 V$ for row (sites) scores 
+ $V^\star = {D_c}^{-1} {P_0}^\top U$ for column (species) scores
:::

### Pre-computations

To check our results, we first perform a dc-CA with ade4:

```{r}
ca <- dudi.coa(Y, 
               nf = c-1, 
               scannf = FALSE)

dcca <- dpcaiv2(dudi = ca, 
                dfR = E,
                dfQ = T_,
                scannf = FALSE, 
                nf = min(k, l))
```

### Center matrices

We define $P$ as the relative counts of $Y$ and $P_0$ as its centered scaled version:

```{r}
P <- Y/sum(Y)

# Initialize P0 matrix
P0 <- matrix(ncol = ncol(Y), nrow = nrow(Y))
colnames(P0) <- colnames(Y)
rownames(P0) <- rownames(Y)

for(i in 1:nrow(Y)) { # For each row
  for (j in 1:ncol(Y)) { # For each column
    # Do the sum
    pi_ <- sum(P[i, ])
    p_j <- sum(P[, j])
    
    # Compute the transformation
    P0[i, j] <- (P[i, j] - (pi_*p_j))/sqrt(pi_*p_j)
  }
}
```

First, we need to center-scale the traits and environment matrices (resp. $\tilde{T}$ and $\tilde{E}$). To do so, we use the occurrences in matrix $P$ as weights:


```{r}
E_wt <- rowSums(P)/nrow(E)
Escaled <- scalewt(E, wt = E_wt, scale = TRUE)

T_wt <- colSums(P)/nrow(T_)
Tscaled <- scalewt(T_, wt = T_wt, scale = TRUE)
```

```{r}
# Check centering
M1 <- matrix(rep(1, nrow(P)), nrow = 1)
all(abs(M1 %*% diag(rowSums(P)) %*% Escaled) < zero)

M1 <- matrix(rep(1, ncol(P)), nrow = 1)
all(abs(M1 %*% diag(colSums(P)) %*% Tscaled) < zero)
```

This is equivalent to scaling "inflated" versions of these matrices matching the occurrence counts in $P$ (below).

$$
\tilde{E} = \frac{E - \bar{E}_{infl}}{\sigma_{Einfl}}
$$

$$
\tilde{T} = \frac{T - \bar{T}_{infl}}{\sigma_{Tinfl}}
$$

```{r}
#| code-fold: true

# Center E -----
pi_ <- rowSums(P)
Escaled2 <- matrix(nrow = nrow(E), ncol = ncol(E))

for(i in 1:ncol(Escaled)) {
  emean <- sum(E[, i]*pi_/sum(P))
  Escaled2[, i] <- (E[, i] - emean)/(sqrt(sum(pi_*(E[, i] - emean)^2)))
}
# This is the same as computing a mean on inflated data matrix Einfl and centering E with these means

# Center T -----
p_j <- colSums(P)
Tscaled2 <- matrix(nrow = nrow(T_), ncol = ncol(T_))
rownames(Tscaled2) <- rownames(T_)
colnames(Tscaled2) <- colnames(T_)

for(j in 1:ncol(Tscaled)) {
  tmean <- sum(T_[, j]*p_j/sum(P))
  Tscaled2[, j] <- (T_[, j] - tmean)/(sqrt(sum(p_j*(T_[, j] - tmean)^2)))
}
```


```{r}
#| code-fold: true
# Check equivalence between the 2 methods
all(abs(Escaled/Escaled2 - 1) < zero)
all(abs(Tscaled/Tscaled2 - 1) < zero)
```

### Rationale: regression

Then, we search scores $u$ and $v$ that maximize the fourth-corner correlation $u^\top P v$ (where $u$ are the sites (rows) scores and $v$ are the species (columns) scores).

In this framework, we define $u$ and $v$ as linear combinations of traits and environmental variables: $u = \tilde{E}b$ and $v = \tilde{T}c$. 

So in the end, we need to maximize $u^\top P_0 v$ with respect to the coefficients vectors $b$ and $c$:

$$
\max_{b, c}(u^\top P v) = \max_{b, c}\left(\left[\tilde{E}b\right]^\top P \tilde{T}c \right) 
$$

::: {.callout-note}
These equations are written for the first axis, but we can also write them in matrix form:

$$
\max_{B, C}(U^\top P V) = \max_{B, C}\left(\left[\tilde{E}B\right]^\top P \tilde{T}C \right) 
$$
::: 

We define the diagonal matrices $D_r$ and $D_c$, which contain the column and row sums of $P$ (respectively). We introduce the following constraint on $U$ and $V$: $u^\top D_r u = 1$ and $v^\top D_c v = 1$. It means that the matrices $U$ and $V$ are orthonormal with respect to the weights $D_r$ and $D_c$. (In fact, these constraints will be relaxed later depending on the scaling (see below)).

To find the coefficients $B$ and $C$ defined above, we can perform a SVD or two diagonalizations (see below).

### SVD

To find the coefficients $B$ and $C$, we perform the SVD of a matrix $D$ defined as:

$$
D = \underbrace{[\tilde{E}^\top D_r \tilde{E}]^{-1/2}}_{L^{-1/2}} \underbrace{\tilde{E}^\top P \tilde{T}}_{R} \underbrace{[\tilde{T}^\top D_c \tilde{T}]^{-1/2}}_{K^{-1/2}}
$$ 

```{r}
# Define weights
dr <- rowSums(P)
dc <- colSums(P)

Dr <- diag(dr)
Dc <- diag(dc)
```

```{r}
# Define intermediate matrices
L <- t(Escaled) %*% Dr %*% Escaled
R <- t(Escaled) %*% P %*% Tscaled
K <- t(Tscaled) %*% Dc %*% Tscaled
```

With our dataset:

```{r}
D <- solve(sqrtm(t(Escaled) %*% Dr %*% Escaled)) %*%
  t(Escaled) %*% P %*% Tscaled %*%
  solve(sqrtm(t(Tscaled) %*% Dc %*% Tscaled))
```

We perform the SVD of $D$: 

$$D = B_0 \Delta C_0^\top$$

```{r}
sv <- svd(D)

# Singular values
delta <- sv$d
Delta <- diag(delta)

# Eigenvalues
lambda <- delta^2
Lambda <- diag(lambda)

# Eigenvectors
B0 <- sv$u
dim(B0) # l x l

C0 <- sv$v
dim(C0) # k x l
```


### Diagonalizations

Equivalently, we can also diagonalize the matrices $M$ and $M_2$ defined below.


$M$ is defined as:

$$
M = [\underbrace{\tilde{E}^\top D_r \tilde{E}}_{L} ]^{-1} \underbrace{\tilde{E}^\top P \tilde{T}}_{R} [ \underbrace{\tilde{T}^\top D_c \tilde{T}}_{K} ]^{-1} \underbrace{\tilde{T}^\top P^\top \tilde{E}}_{R^\top}
$${#eq-M}

We can also view $M$ as:

$$
M = \underbrace{\left[\tilde{E}^\top D_r \tilde{E} \right]^{-1} \tilde{E}^\top P \tilde{T}}_{\hat{E}_{center} = \beta \tilde{T}} \underbrace{\left[\tilde{T}^\top D_c \tilde{T} \right]^{-1} \tilde{T}^\top P^\top \tilde{E}}_{\hat{T}_{center} = \gamma \tilde{E}}
$$

where $\hat{E}_{center}$ is the predicted values of *sites* variables from *species* traits. Reciprocally, $\hat{T}_{center}$ is the predicted values of *species* traits from *sites* variables. $\beta$ and $\gamma$ are the vectors of the regression coefficients. From @eq-M, we have $\beta = \left[\tilde{E}^\top D_r \tilde{E} \right]^{-1} \tilde{E}^\top P$ and $\gamma = \left[\tilde{T}^\top D_c \tilde{T} \right]^{-1} \tilde{T}^\top P^\top$.

```{r}
M <- solve(t(Escaled) %*% Dr %*% Escaled) %*% t(Escaled) %*% P %*% Tscaled %*% solve(t(Tscaled) %*% Dc %*% Tscaled) %*% t(Tscaled) %*% t(P) %*% Escaled
```


$M_2$ is defined as:

$$
M_2 = \underbrace{\left[\tilde{T}^\top D_c \tilde{T} \right]^{-1} \tilde{T}^\top P^\top \tilde{E}}_{\hat{T}_{center} = \gamma \tilde{E}} \underbrace{\left[\tilde{E}^\top D_r \tilde{E} \right]^{-1} \tilde{E}^\top P \tilde{T}}_{\hat{E}_{center} = \beta \tilde{T}}
$$

```{r}
M2 <- solve(t(Tscaled) %*% Dc %*% Tscaled) %*% t(Tscaled) %*% t(P) %*% Escaled %*% solve(t(Escaled) %*% Dr %*% Escaled) %*% t(Escaled) %*% P %*% Tscaled 
```

The eigenvectors matrices of these diagonalizations give us $B$ and $C$:

+ To find $B$, we diagonalize $M$:

$$
M = B \Lambda_b B^{-1}
$$

```{r}
# Diagonalize M
eigB <- eigen(M)
lambdaB <- eigB$values
lambdaB # All l = 6 non-null eigenvectors
B_diag <- eigB$vectors

# Check that the result checks out with SVD
all(abs(abs(B_diag/B0) - 1) < zero) # it isn't equal

apply(B_diag, 2, function(x) sqrt(sum(x^2))) # but B_diag is of norm 1 (like B0)
```

+ Similarly, to find $C$, we must diagonalize $M_2$:

$$
M_2 = C \Lambda_c C^{-1}
$$

```{r}
# Diagonalize M2
eigC <- eigen(M2)
lambdaC <- eigC$values
lambdaC # l = 6 non-null eigenvalues
C_diag <- eigC$vectors

# Check that the result checks out with SVD
all(abs(abs(C_diag[, 1:l]/C0) - 1) < zero) # it isn't equal

apply(C_diag[, 1:l], 2, function(x) sqrt(sum(x^2))) # but C_diag is of norm 1 (like C0)
```


```{r}
# The two diagonalizations have the same non-null eigenvectors
all(lambdaB - lambdaC[1:l] < zero)
```

The eigenvalues of the SVD $\Lambda = \Delta^2$ are the same as $\Lambda_b$ and $\Lambda_c$.

```{r}
all(abs(lambda - lambdaB) < zero)
```

### Compute $B$ and $C$

We also define the following "scalings" for the coefficients:

$$
\left\{
\begin{array}{ll}
  B &= L^{-1/2} B_0\\
  C &= K^{-1/2} C_0
\end{array}
\right.
$$

(We recall that $L = \tilde{E}^\top D_r \tilde{E}$ and $K = \tilde{T}^\top D_c \tilde{T}$). 

```{r}
B <- solve(sqrtm(L)) %*% B0 
C <- solve(sqrtm(K)) %*% C0
```

```{r}
# B0 and C0 are of norm 1
apply(B0, 2, function(x) sqrt(sum(x^2))) 
apply(C0, 2, function(x) sqrt(sum(x^2))) 

# B and C are normed with L and K

# t(B) L B is the identity matrix
id_l <- diag(1, nrow = l, ncol = l)
all(abs((t(B) %*% L %*% B) - id_l) < zero)

# t(C) K C is the identity matrix
id_c <- diag(c(rep(1, k-1), 0),
             nrow = k, ncol = k) # Identity minus last vector (zero)
C_k <- cbind(C, rep(0, k)) # Add null eigenvector
all(abs(t(C_k) %*% K %*% C_k - id_c) < zero)
```


### Sites and species scores

#### LC scores

Using the coefficients $B$ and $C$, we can now define scores for sites and species scores as a linear combination of their variables:

+ sites scores predicted by environmental variables are $U = \tilde{E} B$

```{r}
U <- Escaled %*% B0

# Check it is the same as ade4 scores
all(abs(U/dcca$l1) - 1 < zero)
```


+ species scores predicted by their traits are $V = \tilde{T} C$

```{r}
V <- Tscaled %*% C

# Check it is the same as ade4 scores
all(abs(V/dcca$c1) - 1 < zero)
```


#### WA scores

We can also define species scores as the mean of sites LC scores (and the reverse):

+ sites scores as weighted averages of species scores are $U^\star = D_r^{-1} P V$

```{r}
Ustar <- solve(Dr) %*% P %*% V

# Check it is the same as ade4 scores
all(abs(Ustar/dcca$lsR) - 1 < zero)
```

+ species scores as weighted averages of sites scores are $V^\star = D_c^{-1} P^\top U$

```{r}
Vstar <- solve(Dc) %*% t(P) %*% U

# Check it is the same as ade4 scores
all(abs(Vstar/dcca$lsQ) - 1 < zero)
```


## Scalings

::: {.callout-tip title="*TL;DR*"}


There are two types of coordinates: linear combination scores (LC scores) and weighted averages scores (WA scores) for the sites and species individuals.

**The general formulas are:**

+ the LC scores for rows (sites) are $U_i = \tilde{E} B \Lambda^{\alpha/2}$
+ the LC scores for columns (species) are $V_i = \tilde{T}C_i \Lambda^{(1 - \alpha)/2}$
+ the WA scores for rows (sites) are $U^\star_i = {D_r}^{-1} P V_i$
+ the WA scores for columns (species) are $V^\star_i = {D_c}^{-1} P^\top U_i$
+ the correlations with axes for the rows (environmental variables) are $BS_{Bi} = B  \Lambda^{\alpha/2}$
+ the correlations with axes for the columns (species traits) are $BS_{Ci} = C \Lambda^{(\alpha - 1)/2}$

In these formulas, note that the WA scores for one dimension are computed from the predicted scores of the other dimension.

**Scaling type 1 ($\alpha = 1$)**

+ the LC scores for rows (sites) are $U_1 = \tilde{E} B \Lambda^{1/2}$ (`li`)
+ the LC scores for columns (species) are $V_1 = \tilde{T}C$ (`c1`)
+ the WA scores for rows (sites) are $U^\star_1 = {D_r}^{-1} P V_1$ (`lsR`)
+ the WA scores for columns (species) are $V^\star_1 = {D_c}^{-1} P^\top U_1$
+ the correlations with axes for the rows (environmental variables) are $BS_{B1} = B  \Lambda^{1/2}$
+ the correlations with axes for the columns (species traits) are $BS_{C1} = C$


**Scaling type 2 ($\alpha = 0$)**

+ the LC scores for rows (sites) are $U_2 = \tilde{E} B$ (`l1`)
+ the LC scores for columns (species) are $V_2 = \tilde{T}C \Lambda^{1/2}$ (`co`)
+ the WA scores for rows (sites) are $U^\star_2 = {D_r}^{-1} P V_2$
+ the WA scores for columns (species) are $V^\star_2 = {D_c}^{-1} P^\top U_2$ (`lsQ`)
+ the correlations with axes for the rows (environmental variables) are $BS_{B2} = B$
+ the correlations with axes for the columns (species traits) are $BS_{C2} = C  \Lambda^{1/2}$

**Scaling type 3 ($\alpha = 1/2$)**

+ the LC scores for rows (sites) are $U_3 = \tilde{E} B \Lambda^{1/4}$
+ the LC scores for columns (species) are $V_3 = \tilde{T} C \Lambda^{1/4}$
+ the WA scores for rows (sites) are $U^\star_3 = {D_r}^{-1} P V_3$
+ the WA scores for columns (species) are $V^\star_3 = {D_c}^{-1} P^\top U_3$
+ the correlations with axes for the rows (environmental variables) are $BS_{B3} = B \Lambda^{1/4}$
+ the correlations with axes for the columns (species traits) are $BS_{C3} = C \Lambda^{1/4}$


$\alpha$ changes the interpretation of the correlations vectors: 

+ the following vectors are used for intra-set correlations: $BS_{B2}$ ($\alpha = 0$), to approximate the correlations between the environmental/row variables and the sites, and $BS_{C1}$ ($\alpha = 1$) for the correlations between the traits/column variables and the species.
+ inter-set correlations (fourth-corner): $BS_{B1}$ ($\alpha = 1$) approximates the correlation between environmental/row variables and columns/species and $BS_{C2}$ ($\alpha = 0$) approximates the correlation between columns variables/traits and sites.
+ $BS_{B3}$ and $BS_{C3}$ ($\alpha = 1/2$) is the geometric mean of scalings 1 and 2.

When plotting the correlation circle, to look at the correlations between variables of the same set (traits or environmental variables), we should use $BS_{B2}$ and $BS_{C1}$. There are the scores returned by `ade4`.
:::


### Scaling type 1

This type of scaling preserves the distances between rows ($\alpha = 1$).

+ the rows (sites) scores can be $U^\star_1$ (WA scores) or $U_1$ (LC scores)  
+ the columns (species) scores can be $V^\star_1$ (WA scores) or $V_1$ (LC scores) 
+ rows variables (environmental variables) correlations are $BS_{B1}$.
+ columns variables (species traits) correlations are $BS_{C1}$


With the scaling type 1, $BS_{B1}$ represents the correlation between environmental variables and species and $BS_{C1}$ represents the correlation between species traits and species.

```{r}
# LC scores
U1 <- Escaled %*% B %*% Delta # rows
V1 <- Tscaled %*% C # columns

# WA scores
Ustar1 <- solve(Dr) %*% P %*% V # rows
Vstar1 <- solve(Dc) %*% t(P) %*% U  # columns

# # Variables scores
BS_B1 <- B %*% Delta
BS_C1 <- C

# Normalize
BS_B1norm <- normalize(BS_B1, 1)
BS_C1norm <- normalize(BS_C1, 1)

BS_C1 <- t(Tscaled) %*% Dc %*% V
```

```{r}
# Compare results to ade4
all(abs(U1/dcca$li) - 1 < zero) # sites LC
all(abs(V1/dcca$c1) - 1 < zero) # spp LC

all(abs(Ustar1/dcca$lsR) - 1 < zero) # sites WA
all(abs(abs(Vstar1/dcca$lsQ) - 1) < zero) # not stored in ade4

all(abs(abs(BS_C1/dcca$corQ) - 1) < zero)
```

#### Plots

We can either plot:

+ sites as $U_1$ (`li`) and species as $V_1$ (`c1`) (left)
+ sites as $U^\star_1$ (`lsR`) and species as $V_1$ (`c1`) (right)

```{r, fig.height=6, fig.width=8, warning=FALSE}
#| code-fold: true

# LC scores
glc <- multiplot(indiv_row = U1, # LC for sites
                 indiv_col = V1, # LC1 for spp
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 # var_row = BS_B1norm, var_row_lab = colnames(E),
                 # var_col = BS_C1norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("Sites LC scores with species\nnormed at 1")

# WA scores
gwa <- multiplot(indiv_row = Ustar1, # WA for sites
                 indiv_col = V1, # LC1 for spp
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 # var_row = BS_B1norm, var_row_lab = colnames(E),
                 # var_col = BS_C1norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("Sites WA scores with species\nnormed at 1")

glc + gwa + 
  plot_layout(axis_titles = "collect")

s.corcircle(BS_B1norm,
            labels = colnames(E),
            plabels.col = params$colsite)
s.corcircle(BS_C1norm,
            labels = colnames(T_),
            plabels.col = params$colspp, add = TRUE)
```

On this plot, we can interpret 5 sets of pairs:

+ row and columns variables ($BS_{B1}$ - $BS_{C1}$): angles between arrows of rows and environmental variables represent their fourth-corner correlation. For instance, the correlation between `mass` and `perc_forests` is large (arrows size + direction).
+ species and environmental variables ($BS_{B1}$ - $V$): species points on the right (e.g. sp1) are located in sites with a high `location` variable.
+ sites and species traits ($BS_{C1}$ - $U$): we can understand sites using their community weighted mean (eg site 26 has birds with high mass).
+ species and species traits ($BS_{C1}$ - $U$): for instance, species 12 has large eggs.

The last pair is the individuals on each plot:

+ On the first plot (with sites LC scores), $U_1$ and $V_1$ form a biplot of contingency ratio [@terbraak2018]
+ On the second plot (with sites WA scores), $U^\star_1$ is placed at the mean of $V_1$.

Moreover, species traits ($BS_C^1$): arrows indicate intra-set correlations.

**There is no interpretation for sites - environmental variables ($U_1$ - $BS_{B1}$).**


Same plots with ade4:

```{r}
#| code-fold: true

s.label(dcca$c1, # Species LC1
        plabels.optim = TRUE,
        plabels.col  = params$colspp,
        ppoints.col = params$colspp,
        main = "LC scores for sites")
s.label(dcca$li, # Sites LC
        plabels.optim = TRUE,
        plabels.col  = params$colsite,
        ppoints.col = params$colsite,
        add = TRUE)

s.label(dcca$c1, # Species LC1
        plabels.optim = TRUE,
        plabels.col  = params$colspp,
        ppoints.col = params$colspp,
        main = "WA scores for sites")
s.label(dcca$lsR, # Sites WA scores
        plabels.optim = TRUE,
        plabels.col  = params$colsite,
        ppoints.col = params$colsite,
        add = TRUE)

s.corcircle(dcca$corQ,
            plabels.col = params$colspp)
s.corcircle(dcca$corR,
            plabels.col = params$colsite, add = TRUE)
```


### Scaling type 2

This type of scaling preserves the distances between columns.

+ the rows (sites) scores can be $U^\star_2$ (WA scores) or $U_2$ (LC scores)  
+ the columns (species) scores can be $V^\star_2$ (WA scores) or $V_2$ (LC scores) 
+ rows variables (environmental variables) correlations are $BS_{B2}$.
+ columns variables (species traits) correlations are $BS_{C2}$


With the scaling type 2, $BS_{B2}$ represents the correlation between environmental variables and sites and $BS_{C2}$ represents the correlation between species traits and sites.

```{r}
# LC scores
U2 <- Escaled %*% B # rows
V2 <- Tscaled %*% C %*% Delta # columns

# WA scores
Ustar2 <- solve(Dr) %*% P %*% V2 # rows
Vstar2 <- solve(Dc) %*% t(P) %*% U2 # columns

# Variables scores
BS_B2 <- t(Escaled) %*% Dr %*% U2
# BS_B2 <- B
# BS_C2 <- C %*% Delta
# 
# # Normalize
# BS_B2norm <- normalize(BS_B2, 1)
# BS_C2norm <- normalize(BS_C2, 1)
```

```{r}
all(abs(U2/dcca$l1) - 1 < zero) # LC sites
all(abs(V2/dcca$co) - 1 < zero) # LC spp

all(abs(Vstar2/dcca$lsQ) - 1 < zero) # WA spp
all(abs(Ustar2/dcca$lsR) - 1 < zero) # WA site

all(abs(abs(BS_B2/dcca$corR) - 1) < zero)
```

#### Plots

We can either plot:

+ sites as $U_2$ (`l1`) and species as $V_2$ (`co`) (left)
+ sites as $U_2$ (`l1`) and species as $V^\star_2$ (`lsQ`) (right)

```{r, fig.height=6, fig.width=8, warning=FALSE}
#| code-fold: true

# LC scores
glc <- multiplot(indiv_row = U2, indiv_col = V2, 
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 # var_row = BS_B2norm, var_row_lab = colnames(E),
                 # var_col = BS_C2norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("Species LC scores with sites\nnormed at 1")

# WA scores
gwa <- multiplot(indiv_row = U2, indiv_col = Vstar2, 
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 # var_row = BS_B2norm, var_row_lab = colnames(E),
                 # var_col = BS_C2norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("Species WA scores with sites\nnormed at 1")

glc + gwa + 
  plot_layout(axis_titles = "collect")

# s.corcircle(BS_B2norm,
#             labels = colnames(E),
#             plabels.col = params$colsite)
# s.corcircle(BS_C2norm,
#             labels = colnames(T_),
#             plabels.col = params$colspp, add = TRUE)
```


**There is no interpretation for species - species traits ($V_2$ - $BS_{C2}$).**


With ade4:

```{r}
#| code-fold: true

s.label(dcca$li, # LC1 sites
        plabels.optim = TRUE,
        plabels.col  = params$colsite,
        ppoints.col = params$colsite,
        main = "LC scores for species")
s.label(dcca$co, # LC species
        plabels.optim = TRUE,
        plabels.col  = params$colspp,
        ppoints.col = params$colspp,
        add = TRUE)

s.label(dcca$l1, # LC1 sites
        plabels.optim = TRUE,
        plabels.col  = params$colsite,
        ppoints.col = params$colsite,
        main = "WA scores for species")
s.label(dcca$lsQ, # WA species
        plabels.optim = TRUE,
        plabels.col  = params$colspp,
        ppoints.col = params$colspp,
        add = TRUE)
```


### Scaling type 3

This type of scaling is an intermediate between scalings 1 and 2.

+ the rows (sites) scores can be $U^\star_3$ (WA scores) or $U_3$ (LC scores)  
+ the columns (species) scores can be $V^\star_3$ (WA scores) or $V_3$ (LC scores) 
+ rows variables (environmental variables) correlations are $BS_{B3}$.
+ columns variables (species traits) correlations are $BS_{C3}$

With the scaling type 3, $BS_{B3}$ and $BS_{C3}$ represent the geometric mean of their correlation with species and sites.

```{r}
# LC scores
U3 <- Escaled %*% B %*% Delta^(1/2) # rows
V3 <- Tscaled %*% C %*% Delta^(1/2) # columns

# WA scores
Ustar3 <- solve(Dr) %*% P %*% V3 # rows
Vstar3 <- solve(Dc) %*% t(P) %*% U3 # columns

# Variables scores
BS_B3 <- B %*% Delta^(1/2)
BS_C3 <- C %*% Delta^(1/2)

# Normalize
BS_B3norm <- normalize(BS_B3, 1)
BS_C3norm <- normalize(BS_C3, 1)
```

#### Plots

```{r, fig.height=6, fig.width=6, warning=FALSE}
#| code-fold: true

# LC scores
glc <- multiplot(indiv_row = U3, indiv_col = V3, 
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 # var_row = BS_B3norm, var_row_lab = colnames(E),
                 # var_col = BS_C3norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("LC scores")

# WA scores
gwa <- multiplot(indiv_row = Ustar3, indiv_col = Vstar3, 
                 indiv_row_lab = rownames(Y), indiv_col_lab = colnames(Y), 
                 # var_row = BS_B3norm, var_row_lab = colnames(E),
                 # var_col = BS_C3norm, var_col_lab = colnames(T_),
                 row_color = params$colsite, col_color = params$colspp,
                 eig = lambda) +
  ggtitle("WA scores")

glc + gwa + 
  plot_layout(axis_titles = "collect")

s.corcircle(BS_B3norm,
            labels = colnames(E),
            plabels.col = params$colsite)
s.corcircle(BS_C3norm,
            labels = colnames(T_),
            plabels.col = params$colspp, add = TRUE)
```

## Interpretation

This method finds the linear correlation of row explanatory variables (environmental variables) and the linear correlation of columns explanatory variables (species traits) that maximizes the fourth-corner correlation, i.e. the correlation between these linear combinations of row and columns-variables.

There are other related methods, that have been better described and also more used in ecology: RLQ, community weighted means RDA (CMW-RDA). 

Contrary to RLQ, dc-CA takes into account the correlation between the row and column variables. Thus, while RLQ can analyze any number of row and column variables, it is not the case with dc-CA the number of row and column variables must not be large compared to the number of rows/columns in the tables. Also, dc-CA maximizes correlation and RLQ maximizes covariance [@terbraak2018].

The eigenvalues of dc-CA are the squares of the fourth-corner correlations.