---
title: "Double-constrained Correspondence Analysis (dc-CA)"
params:
  colsite: "burlywood4"
  colspp: "cornflowerblue"
---


```{r, results='hide', message=FALSE}
#| code-fold: true

# Paths
library(here)

# Multivariate analysis
library(ade4)
library(adegraphics)

# dc-CA
source(here("functions/dpcaiv2-ade4.R"))

# Matrix algebra
library(expm)

# Plots
library(ggplot2)
source(here("functions/plot.R"))
```


*The contents of this page relies heavily on [@terbraak2018].*

## Introduction

dc-CA was developed as a natural extension of CCA and has been used to study the relationship between species traits and environmental variables.

In dc-CA, we have 3 matrices:

+ A data matrix $Y$ ($r \times c$)
+ A matrix of predictor variables $E$ ($r \times l$)
+ A matrix of predictor variables $T$ ($c \times k$)


```{r, echo = FALSE, message = FALSE}
dat <- readRDS(here("data/Barbaro2012.rds"))
Y <- dat$comm
E <- dat$envir
T_ <- dat$traits

r <- dim(Y)[1]
c <- dim(Y)[2]
l <- dim(E)[2]
k <- dim(T_)[2]

plotmat(r = r, c = c, 
        E = TRUE, T_ = TRUE,
        l = l, k = k)
```

The aim of dc-CA is to find a linear combination of the predictor variables in $E$ and $T$ (environmental variables and traits) that maximizes the correlation.

Below are these matrices for our data:

$Y =$
```{r, echo = FALSE, message = FALSE}
knitr::kable(Y)
```


$E =$
```{r, echo = FALSE, message = FALSE}
knitr::kable(E)
```

$T =$
```{r, echo = FALSE, message = FALSE}
knitr::kable(T_)
```

```{r}
(r <- dim(Y)[1])
(c <- dim(Y)[2])
(l <- dim(E)[2])
(k <- dim(T_)[2])
```

dc-CA must not have to many traits compared to species: that is a disadvantage compared to RLQ, but on the other hand dc-CA allows to see relationships that RLQ would miss [@terbraak2018].

There are several ways to perform dc-CA [@terbraak2018], notably:

-   singular value decomposition (the method used here)
-   an iterative method *à la* reciprocal averaging
-   canonical correlation analysis between $T$ et $E$, weighted by $Y$

## Computation

::: {.callout-tip title="*TL;DR*"}
We perform the SVD of $D = [E_{center}' D(y_{i \cdot}) E_{center}]^{-1/2} E_{center}' Y T_{center} [T_{center}' D(y_{\cdot j}) T_{center}]^{-1/2}$

$$
D = P \Lambda_{SDV} Q'
$$
This allows us to find the eigenvectors $P$ (rows or environment eigenvectors) and $Q$ (columns or traits eigenvectors).

The eigenvalues of the dc-CA are the squared eigenvalues of the SVD: $\Lambda_{dc-CA} = \Lambda_{SVD}^2$

There are $\min(k, l)$ non-null eigenvalues???

Then, the regression coefficients for explanatory variables can be found with 

+ $B = [E_{center}'D(y_{i\cdot})E_{center}]^{-1/2}P$ (row explanatory variables = environmental variables)
+ $C = [T_{center}' D(y_{\cdot j})T_{center}]^{-1/2}Q$ (columns explanatory variables = species traits)

The rows/column scores are detailed below in the Scalings section.
:::


First, we need to center the traits and environment matrices (resp. $T_{cent}$ and $E_{cent}$). To do that, we have to compute "inflated" versions of these matrices matching the occurrence counts in $Y$.

$$
E_{stand} = E - \bar{E}_{infl} =\left[e_i \sum_i y_{i\cdot}e_i/y_{\cdot\cdot} \right]
$$

$$
T_{stand} = T - \bar{T}_{infl} = \left[t_j \sum_j y_{\cdot j}t_j/y_{\cdot\cdot} \right]
$$

With our data:

```{r}
# Center E -----
yi_ <- rowSums(Y)
Ecenter <- matrix(nrow = nrow(E), ncol = ncol(E))

for(i in 1:ncol(Ecenter)) {
  Ecenter[, i] <- E[, i] - sum(E[, i]*yi_)/sum(Y)
}
# This is the same as computing a mean on inflated data matrix Einfl and centering E with these means

# Center T -----
y_j <- colSums(Y)
Tcenter <- matrix(nrow = nrow(T_), ncol = ncol(T_))
rownames(Tcenter) <- rownames(T_)
colnames(Tcenter) <- colnames(T_)

for(j in 1:ncol(Tcenter)) {
  Tcenter[, j] <- T_[, j] - (sum(T_[, j]*y_j)/sum(Y))
}
```

```{r}
# Check centering -----
M1 <- matrix(rep(1, nrow(Y)), nrow = 1)
all((M1 %*% diag(rowSums(Y)) %*% Ecenter) < 10e-10)

M1 <- matrix(rep(1, ncol(Y)), nrow = 1)
all((M1 %*% diag(colSums(Y)) %*% Tcenter) < 10e-10)
```

Maximizing the fourth-corner correlation means finding $\hat{u}$ and $u$ that maximize $\hat{u}'Yu$ (where $\hat{u}$ are the sites (rows) scores and $u$ are the species (columns) scores).

$\hat{u}$ and $u$ are linear combinations of traits and environmental variables: $\hat{u} = E_{center}b$ and $u = T_{center}c$. 

So in the end, we need to maximize $\hat{u}'Yu$ with respect to the coefficients vectors $b$ and $c$:

$$
\max_{b, c}(\hat{u}'Yu) = \max_{b, c}\left(\left[E_{center}b\right]'Y  T_{center}c \right) 
$$

::: {.callout-note}
These equations are written for the first axis, but we can also write them in matrix form:

$$
\max_{B, C}(\hat{U}'YU) = \max_{B, C}\left(\left[E_{center}B\right]'Y T_{center}C \right) 
$$
::: 

In addition, we introduce the following constraint on the norm of the columns vectors of $\hat{U}$ and $U$: $\hat{u}' D(y_{i\cdot}) \hat{u} = 1$ and $u' D(y_{\cdot j})u = 1$. In fact, tjese constraints will be relaxed later depending on the scaling (see below).

To find the coefficients $B$ and $C$ defined above, we need to diagonalize the following matrices $M$ and $M_2$:

+ To find $B$, we must diagonalize $M$:
$$
M = \left[E_{center}' D(y_{i\cdot}) E_{center} \right]^{-1} E_{center}'YT_{center} \left[T_{center}' D(y_{\cdot j}) T_{center} \right]^{-1} T_{center}' Y' E_{center}
$$
Where matrices $D(y_{i \cdot})$ and $D(y_{\cdot j})$ are the diagonal matrices with the column and row sums (respectively). They are analogous to matrices $D(p_{i \cdot})$ and $D(p_{\cdot j})$ defined for CA and CCA (but we use $Y$ instead of $P$ to define the margins).
We can view $M$ as:
$$
M = \hat{E}_{center} \hat{T}_{center} = \beta T_{center} \gamma E_{center}
$$
With $\beta = \left[E_{center}' D(y_{i\cdot}) E_{center} \right]^{-1} E_{center}'Y$ and $\gamma = \left[T_{center}' D(y_{\cdot j}) T_{center} \right]^{-1} T_{center}' Y'$. Here, we predict the environment with traits so that $\hat{E}_{center} = \beta T_{center}$ and the traits with the environment so that $\hat{T}_{center} = \gamma E_{center}$.
So we maximize the correlation between $\hat{E}_{center}$ and $\hat{T}_{center}$.

+ Similarly, to find $C$, we must diagonalize $M_2$:

$$
M_2 = \left[T_{center}' D(y_{\cdot j}) T_{center} \right]^{-1} T_{center}' Y' E_{center} \left[E_{center}' D(y_{i\cdot}) E_{center} \right]^{-1} E_{center}'YT_{center}
$$
Here, we can view $M_2$ as:
$$$M_2 = \hat{T}_{center} \hat{E}_{center} = \gamma E_{center} \beta T_{center}$$

The eigenvectors matrices of these diagonalizations give us $B$ and $C$:

$$
M = B \Lambda_b B^{-1} ~~ \text{and} ~~ M_2 = C \Lambda_c C^{-1}
$$

We can either diagonalize $M$ and $M_2$ of perform a single SVD of a matrix $D$ defined below and get a similar result. Both methods are presented below.

### Diagonalizations

We diagonalize $M$ and $M_2$:

```{r}
# Define weights
Dyi_ <- diag(rowSums(Y))
Dy_j <- diag(colSums(Y))

# Compute M
M <- solve(t(Ecenter) %*% Dyi_ %*% Ecenter) %*% t(Ecenter) %*% Y %*% Tcenter %*% solve(t(Tcenter) %*% Dy_j %*% Tcenter) %*% t(Tcenter) %*% t(Y) %*% Ecenter

# Compute M2
M2 <- solve(t(Tcenter) %*% Dy_j %*% Tcenter) %*% t(Tcenter) %*% t(Y) %*% Ecenter %*% solve(t(Ecenter) %*% Dyi_ %*% Ecenter) %*% t(Ecenter) %*% Y %*% Tcenter 
```


```{r}
# Diagonalize M
eigB <- eigen(M)
lambdaB <- eigB$values
lambdaB # All non-null eigenvectors
vB <- eigB$vectors

# Diagonalize M2
eigC <- eigen(M2)
lambdaC <- eigC$values
lambdaC # six non-null eigenvalues
vC <- eigC$vectors

all(lambdaB - lambdaC[1:l] < 10e-10)
```


  ### SVD

Alternatively, we can compute the SVD of a matrix $D$ computed from $M$ and get a similar result. $D$ is defined as:

$$
D = M^{1/2} = [E_{center}' D(y_{i \cdot}) E_{center}]^{-1/2} E_{center}' Y T_{center} [T_{center}' D(y_{\cdot j}) T_{center}]^{-1/2}
$$ 

With our dataset:


```{r}
D <- solve(sqrtm(t(Ecenter) %*% Dyi_ %*% Ecenter)) %*%
  t(Ecenter) %*% Y %*% Tcenter %*%
  solve(sqrtm(t(Tcenter) %*% Dy_j %*% Tcenter))
```

We perform the SVD of $D$: $D = P \Lambda_{SVD} Q'$.

```{r}
sv <- svd(D)

lambda_svd <- sv$d
Lambda_svd <- diag(lambda_svd)

P_svd <- sv$u
Q_svd <- sv$v

dim(P_svd)
dim(Q_svd)
```


The eigenvalues of the SVD are the square roots of the final eigenvalues obtained by diagonalization: $\Lambda_{dc-CA} = \Lambda_{SVD}^2$

```{r}
lambda <- lambda_svd^2
Lambda <- diag(lambda)
```

Now, we can compute the regression coefficients of the sites environmental variables ($B$) as:

$$
B = [E_{center}'D(y_{i\cdot})E_{center}]^{-1/2}P \Lambda_{SVD}
$$

```{r}
B <- solve(sqrtm(t(Ecenter) %*% Dyi_ %*% Ecenter)) %*%  P_svd %*% Lambda_svd
```

Same for the coefficients of species traits $C$:

$$
C = [T_{center}' D(y_{\cdot j})T_{center}]^{-1/2}Q\Lambda_{SVD}
$$

```{r}
C <- solve(sqrtm(t(Tcenter) %*% Dy_j %*% Tcenter)) %*% Q_svd %*% Lambda_svd
```

```{r}
# Compare both methods
B/vB # The eigenvectors are almost the same
apply(B, 2, function(x) sqrt(sum(x^2))) # column vectors of B have different norms
apply(vB, 2, function(x) sqrt(sum(x^2))) # column vectors of vB are of norm 1

C/vC[, 1:l] # The eigenvectors are almost the same
apply(C, 2, function(x) sqrt(sum(x^2))) # column vectors of C have different norms
apply(vC, 2, function(x) sqrt(sum(x^2))) # column vectors of vC are of norm 1
```


## Scalings

::: {.callout-tip title="*TL;DR*"}
There are two types of coordinates: linear combination scores (LC scores) and weighted averages scores (WA scores).

LC scores are computed from the regression coefficients and the predictor variable table associated with the same dimension.

+ Scaling type 1 (that preserve the rows distances): $\hat{Z}_1 = E_{center}B$ (for rows = sites) and $Z_1 = T_{center}C\Lambda_{dc-CA}^{-1/2}$ (for columns = species)
- Scaling type 2 (that preserve the columns distances) : $\hat{Z}_2 = E_{center}B\Lambda_{dc-CA}^{-1/2}$ (for rows = sites) and  $Z_2 = T_{center}C$ (for columns = species)
- Scaling type 3 (symmetric scaling): $\hat{Z}_3 =  E_{center}B\Lambda_{dc-CA}^{-1/4}$ (for rows = sites) and  $Z_3 = T_{center}C\Lambda_{dc-CA}^{-1/4}$ (for columns = species)

***Note: these results are in disagreement with the article of [@terbraak2018], since the scalings of the columns should be times $\Lambda^{-1}$, and there is a problem with the variance of `$co`. We find that there should be $\Lambda^{1-\alpha}$ in (17) instead of $\Lambda^{\alpha-1}$.***

WA scores are computed from the predicted coordinates of the other dimension:

+ The WA row (sites) scores are computed from the $Z_i$ column (species) scores as $\hat{U}_i = D(y_{\cdot i})^{-1} Y Z_i$
+ The WA column (species) scores are computed from the $\hat{Z}_i$ row (sites) scores as $U_i = D(y_{j \cdot})^{-1} Y' \hat{Z}_i$
:::

### Linear combinations (LC) scores

Using the coefficients $B$ and $C$, we can get the *predicted* species/sites scores as a linear combination of coefficients and traits/environment matrices.

```{r}
ca <- dudi.coa(Y, 
               nf = c-1, 
               scannf = FALSE)

dcca <- dpcaiv2(dudi = ca, 
                dfR = E,
                dfQ = T_,
                scannf = FALSE, 
                nf = min(k, l))
```

Coordinates computed with our method are equal to those computed by `ade4`, but we need to multiply them by a scaling factor $\sqrt{y_{\cdot \cdot}}$.

```{r}
(scaling <- sqrt(sum(Y)))
```

```{r, echo=FALSE}
# Compare ade4 and computed coefficients
B*scaling/dcca$faR[-1,]
C*scaling/dcca$faQ[-1,]
```

- Scaling type 1 (that preserve the rows distances): $\hat{Z}_1 = E_{center}B$ (for rows = sites) and  $Z_1 = T_{center}C \Lambda_{SVD}^{-1} = T_{center}C \Lambda_{dc-CA}^{-1/2}$ (for columns = species).

```{r}
Zhat1 <- Ecenter %*% B
Z1 <- Tcenter %*% C %*% diag(lambda^(-1/2))
```


```{r, include=FALSE}
Zhat1*scaling/dcca$li
Z1*scaling/dcca$c1
```


```{r, include = FALSE}
s.label(dcca$c1, # Species constrained at variance 1
        plabels.optim = TRUE,
        plabels.col  = params$colspp,
        ppoints.col = params$colspp)
s.label(dcca$li, # Sites position by averaging
        plabels.optim = TRUE,
        plabels.col  = params$colsite,
        ppoints.col = params$colsite,
        add = TRUE)
```

```{r, fig.height=6, fig.width=6}
#| code-fold: true
Z1inv <- Z1*scaling
Z1inv[, 1] <- -Z1inv[, 1]

Zhat1inv <- Zhat1*scaling
Zhat1inv[, 1] <- -Zhat1inv[, 1]

plot(x = Z1inv[, 1], y = Z1inv[, 2],
     col = NULL,
     asp = 1,
     main = "LC scores (scaling type 1: row-preserving)")
text(x = Z1inv[, 1], y = Z1inv[, 2],
     label = colnames(Y),
     col = params$colspp,
     asp = 1)
text(x = Zhat1inv[, 1], y = Zhat1inv[, 2],
     label = rownames(Y),
     col = params$colsite,
     asp = 1)
# arrows(x0 = rep(0, nrow(BS1)),
#        y0 = rep(0, nrow(BS1)),
#        x1 = BS1[, 1], 
#        y1 = BS1[, 2],
#        length = 0.1,
#        asp = 1)
# text(x = BS1[, 1], 
#      y = BS1[, 2],
#      label = colnames(E),
#      asp = 1)

abline(h = 0)
abline(v = 0)
```


- Scaling type 2 (that preserve the columns distances) : $\hat{Z}_2 = E_{center}B\Lambda_{SVD}^{-1} = E_{center}B\Lambda_{dc-CA}^{-1/2}$ (for rows = sites) and  $Z_2 = T_{center}C$ (for columns = species) (Here we use the formula that works, different from ter Braak, for Z2 agree)

```{r}
Zhat2 <- Ecenter %*% B %*% diag(lambda_svd^(-1))

Z2 <- Tcenter %*% C %*% diag(lambda_svd^(-2))
Z2_agree <- Tcenter %*% C
```


```{r}
Zhat2*scaling/dcca$l1 # Ok

Z2*scaling/dcca$co # Not ok (but using tB formula)
Z2_agree*scaling/dcca$co # Ok
```

```{r}
# Since ter Braak and ade4 disagree, check the variance of co and Z2 or Z2_agree (it should be the eigenvalues of each axis)

#' Get the variance of vector x
#'
#' @param x The vector
#' @param w The weights
#'
#' @return The variance
varvec <- function(x, w) {
  sum(w*(x-mean(x))^2)
}

res_ade4 <- apply(as.matrix(dcca$co), 
                  2, 
                  function(x) varvec(x, dcca$cw))
res_ade4/dcca$eig
  
res_tB <- apply(Z2*scaling, 
                2, 
                function(x) varvec(x, dcca$cw))
res_tB/dcca$eig 

res_agree <- apply(Z2_agree*scaling, 
                   2, 
                   function(x) varvec(x, dcca$cw))
res_agree/dcca$eig
```

```{r, include = FALSE}
s.label(dcca$l1, # Sites constrained at variance 1
        plabels.optim = TRUE,
        plabels.col  = params$colsite,
        ppoints.col = params$colsite)
s.label(dcca$co, # Species position by averaging
        plabels.optim = TRUE,
        plabels.col  = params$colspp,
        ppoints.col = params$colspp,
        add = TRUE)
```

```{r, fig.height=6, fig.width=6}
#| code-fold: true
Z2inv <- Z2_agree*scaling
Z2inv[, 1] <- -Z2inv[, 1]

Zhat2inv <- Zhat2*scaling
Zhat2inv[, 1] <- -Zhat2inv[, 1]

plot(x = Zhat2inv[, 1], y = Zhat2inv[, 2],
     col = NULL,
     asp = 2,
     main = "LC scores (scaling type 2: column-preserving)")
text(x = Z2inv[, 1], y = Z2inv[, 2],
     label = colnames(Y),
     col = params$colspp,
     asp = 2)
text(x = Zhat2inv[, 1], y = Zhat2inv[, 2],
     label = rownames(Y),
     col = params$colsite,
     asp = 2)
# arrows(x0 = rep(0, nrow(BS2)),
#        y0 = rep(0, nrow(BS2)),
#        x2 = BS2[, 1], 
#        y2 = BS2[, 2],
#        length = 0.2,
#        asp = 2)
# text(x = BS2[, 1], 
#      y = BS2[, 2],
#      label = colnames(E),
#      asp = 2)

abline(h = 0)
abline(v = 0)
```

- Scaling type 3 (symmetric scaling): $\hat{Z}_3 = E_{center}B\Lambda_{SVD}^{-1/2} = E_{center}B\Lambda_{dc-CA}^{-1/4}$ (for rows = sites) and  $Z_3 = T_{center}C\Lambda_{SVD}^{-1/2} = T_{center}C\Lambda_{dc-CA}^{-1/4}$ (for columns = species) (we the heuristic formula)

```{r}
Zhat3 <- Ecenter %*% B %*% diag(lambda^(-1/4))

Z3 <- Tcenter %*% C %*% diag(lambda^(-1/4))
# Should be
# Z3 <- Tcenter %*% C %*% diag(lambda^(-3/4))
```

```{r, fig.height=6, fig.width=6}
#| code-fold: true
Z3inv <- Z3*scaling
Z3inv[, 1] <- -Z3inv[, 1]

Zhat3inv <- Zhat3*scaling
Zhat3inv[, 1] <- -Zhat3inv[, 1]

plot(x = Zhat3inv[, 1], y = Zhat3inv[, 2],
     col = NULL,
     asp = 2,
     xlim = c(-2, 2),
     ylim = c(-2, 2),
     main = "LC scores (scaling type 3: symmetric)")
text(x = Z3inv[, 1], y = Z3inv[, 2],
     label = colnames(Y),
     col = params$colspp,
     asp = 2)
text(x = Zhat3inv[, 1], y = Zhat3inv[, 2],
     label = rownames(Y),
     col = params$colsite,
     asp = 2)
# arrows(x0 = rep(0, nrow(BS2)),
#        y0 = rep(0, nrow(BS2)),
#        x2 = BS2[, 1], 
#        y2 = BS2[, 2],
#        length = 0.2,
#        asp = 2)
# text(x = BS2[, 1], 
#      y = BS2[, 2],
#      label = colnames(E),
#      asp = 2)

abline(h = 0)
abline(v = 0)
```

### Weighted averages (WA) scores

We can also get the weighted averages (WA) scores of the rows/columns. Indeed, rows scores are related to columns scores and conversely.

+ The WA scores for the rows (sites) are $\hat{U}$ and are computed from the ***column (species)*** scores:

$$
\hat{U}_i = D(y_{\cdot i})^{-1} Y Z_i
$$

+ The WA scores for the column (species) are $U$ and are computed from the ***row (sites)*** scores:

$$
U_i = D(y_{j \cdot})^{-1} Y' \hat{Z}_i
$$

```{r}
# Column (species) scores
U1 <- solve(Dy_j) %*% t(Y) %*% Zhat1  
U2 <- solve(Dy_j) %*% t(Y) %*% Zhat2
U3 <- solve(Dy_j) %*% t(Y) %*% Zhat3

# Rows (sites) scores
Uhat1 <- solve(Dyi_) %*% Y %*% Z1
Uhat2 <- solve(Dyi_) %*% Y %*% Z2
Uhat3 <- solve(Dyi_) %*% Y %*% Z3
```


```{r, include=FALSE}
# WA scores
Uhat1*scaling/dcca$lsR
U2*scaling/dcca$lsQ
```


```{r, include = FALSE}
s.label(dcca$lsR, # sites WA scores
        plabels.optim = TRUE,
        plabels.col  = params$colsite,
        ppoints.col = params$colsite,
        xlim = c(-2, 2),
        ylim = c(-2, 2))
s.label(dcca$lsQ, # species WA scores
        plabels.optim = TRUE,
        plabels.col  = params$colspp,
        ppoints.col = params$colspp,
        add = TRUE)
```

```{r, fig.height=6, fig.width=6}
#| code-fold: true
U2inv <- U2*scaling
U2inv[, 1] <- -U2inv[, 1]

Uhat1inv <- Uhat1*scaling
Uhat1inv[, 1] <- -Uhat1inv[, 1]

plot(x = Uhat1inv[, 1], y = Uhat1inv[, 2],
     col = NULL,
     asp = 2,
     xlim = c(-2, 2),
     ylim = c(-2, 2),
     main = "WA scores (with scaling type 1: row-preserving)")
text(x = U2inv[, 1], y = U2inv[, 2],
     label = colnames(Y),
     col = params$colspp,
     asp = 2)
text(x = Uhat1inv[, 1], y = Uhat1inv[, 2],
     label = rownames(Y),
     col = params$colsite,
     asp = 2)
# arrows(x0 = rep(0, nrow(BS2)),
#        y0 = rep(0, nrow(BS2)),
#        x2 = BS2[, 1], 
#        y2 = BS2[, 2],
#        length = 0.2,
#        asp = 2)
# text(x = BS2[, 1], 
#      y = BS2[, 2],
#      label = colnames(E),
#      asp = 2)

abline(h = 0)
abline(v = 0)
```
## More tests

Below, we test the variances of various scores.

+ LC scores for rows:
```{r}
# l1 variances should be 1
apply(as.matrix(dcca$l1), 
      2, 
      function(x) varvec(x, dcca$lw))

# li variances should be eigenvalues
apply(as.matrix(dcca$li), 
      2, 
      function(x) varvec(x, dcca$lw))/dcca$eig
```
+ LC scores for columns
```{r}
# c1 variances should be 1
apply(as.matrix(dcca$c1), 
      2, 
      function(x) varvec(x, dcca$cw))

# co variances should be eigenvalues
apply(as.matrix(dcca$co), 
      2, 
      function(x) varvec(x, dcca$cw))/dcca$eig 
```
```{r}
# Z2 variances should be the eigenvalues
apply(Z2*scaling, 
      2, 
      function(x) varvec(x, dcca$cw))/dcca$eig
```
+ WA scores:
```{r}
# Hypothesis: these scires shound be the eigenvalues
apply(as.matrix(dcca$lsR), 
      2, 
      function(x) varvec(x, dcca$lw))/dcca$eig

# co variances should be eigenvalues
apply(as.matrix(dcca$lsQ), 
      2, 
      function(x) varvec(x, dcca$cw))/dcca$eig
```

## Interpretation

This method finds the linear correlation of row explanatory variables (environmental variables) and the linear correlation of columns explanatory variables (species traits) that maximizes the fourth-corner correlation, i.e. the correlation between these linear combinations of row and columns-variables.

There are other related methods, that have been better described and also more used in ecology: RLQ, community weighted means RDA (CMW-RDA). 

Contrary to RLQ, dc-CA takes into account the correlation between the row and column variables. Thus, while RLQ can analyze any number of row and column variables, it is not the case with dc-CA the number of row and column variables must not be large compared to the number of rows/columns in the tables.

The eigenvaines of dc-CA are the squares of the fourth-corner correlations.