---
title: "Introduction"
bibliography: references.bib
---

This website aims at summarizing the theory behind different methods to analyze count data:

-   correspondence analysis (CA)
-   canonical CA (CCA)
-   double-constrained CA (dc-CA)

It also describes reciprocal scaling [@thioulouse1992] (and its extensions with CCA and dcCA). Reciprocal scaling is a method to compute and visualize the dispersion among count tables analyzed with multivariate methods.

## Dataset

Throughout this document, we will use the data from [@barbaro2012] on bird species distribution in New Zealand extracted from the CESTES database [@jeliazkov2020].

Throughout this document, we will analyze a matrix $Y$ ($r \times c$) and two associated matrices $E$ ($r \times e$) and $T$ ($t \times c$). In our example:

```{r, echo = FALSE, message = FALSE}
library(here)

dat <- readRDS(here("data/Barbaro2012.rds"))
Y <- dat$comm
E <- dat$envir
T_ <- dat$traits

r <- dim(Y)[1]
c <- dim(Y)[2]
l <- dim(E)[2]
k <- dim(T_)[2]
```

-   $Y$ ($r =$ `r r` $\times$ $c =$ `r c`) represents bird species counts in different sites
-   $E$ ($r =$ `r r` $\times$ $l =$ `r l`) represents environmental variables in different sites
-   $T$ ($c =$ `r c` $\times$ $k =$ `r k`) represents bird species traits

```{r, echo = FALSE}
library(ggplot2)

source(here("functions/plot.R"))

plotmat(T_ = TRUE, E = TRUE,
        r = r, c = c, l = l, k = k)
```

