---
title: "Reciprocal scaling"
params:
  colsite: "burlywood4"
  colspp: "cornflowerblue"
---

```{r, results='hide', message=FALSE}
#| code-fold: true

# Paths
library(here)

# Multivariate analysis
library(ade4)
library(adegraphics)

# dc-CA
source(here("functions/dpcaiv2-ade4.R"))

# Reciprocal scaling
source(here("functions/reciprocal.R"))

# Matrix algebra
library(expm)

# Plots
source(here("functions/plot.R"))
library(gridExtra)
```

## Introduction

Reciprocal scaling was introduced by @thioulouse1992. It is a technique allowing to compute the coordinates of each row/column cell in the multivariate space, which then allows to get the conditional mean and variance per row or column.

This technique was initially defined in the frame of correspondence analysis, but it is extended to canonical CA in [reciprocal scaling (CCA)](recscal_cca.qmd) and double-constrained CA in in [reciprocal scaling (dcCA)](recscal_dcca.qmd).

Here, we will analyze matrix $Y$:

```{r, echo = FALSE, message = FALSE}
dat <- readRDS(here("data/Barbaro2012.rds"))
Y <- dat$comm

r <- dim(Y)[1]
c <- dim(Y)[2]

plotmat(r = r, c = c)
```

```{r}
(r <- dim(Y)[1])
(c <- dim(Y)[2])
```

## Computation

::: {.callout-tip title="*TL;DR*"}
Reciprocal scaling gives a score per correspondence (row/column pair) on each ordination axis (instead of a score per row and per column in CA).

These scores can be presented in a matrix $H$ ($\omega \times K$). Here, $\omega = \sum_Y y_i \neq 0$ (number of nonzero cells in $Y$, also called correspondences) and $K$ is the number of eigenvalues (axes) of the CA. Each row of $H$ corresponds to a row/column pair.

**From CA scores**

$H$ can be computed from the scores of the correspondence analysis of $Y$. The score $H_k(i, j)$ for row $i$, column $j$ and axis $k$ is:

$$
H_k(i, j) = \frac{L_k(i) + C_k(j)}{\sqrt{2 \lambda_k \mu_k}}
$$

if $y_{ij} = 0$ then the score $H_k(i, j)$ is not defined.

**From canonical correlation analysis**

We transform the table $Y$ in two tables:

-   $R$ ($\omega \times r$)
-   $C$ ($\omega \times c$)

Each row of $R$ and $C$ contains an indicator vector. The 1 is in the column corresponding to the row (for $R$) or column (for $C$) of the nonzero cell in $Y$.

Additionally, we have a weights vector $w = y_{ij}$ (the same for $R$ and $C$).

Then, we perform a (weighted) canonical correlation analysis of $R_{scaled}$ and $C_{scaled}$. $X_{scaled}$ corresponds to $X$ center and scaled (here, with the weights $w$).

Canonical correlation analysis gives the vectors of canonical coefficients $\rho$ and $\gamma$ (resp. for $R_{scaled}$ and $C_{scaled}$). The scores of the matrices are then $S_R = R_{scaled} \rho$ and $S_C = C_{scaled} \gamma$.

Finally, $H$ is computed as:

$$
H = (S_R + S_C)_{scaled}
$$
:::

The coordinates of row (site) $i$ and column (species) $j$ in the multivariate space on axis $k$ $H_k(i, j)$ can be computed in 2 ways:

-   from the coordinates of the CA
-   from the canonical correlation analysis

### From CA

We perform the CA of table $Y$ ($r \times c$).

```{r}
Ydf <- as.data.frame(Y)
ca <- dudi.coa(Ydf, 
               scannf = FALSE,
               nf = min(r - 1, c - 1))

L <- ca$li
C <- ca$co

lambda <- ca$eig
```

We compute the reciprocal scaling score for comparison:

```{r}
rec_ca <- reciprocal.coa(ca)
```

We can get the coordinates of row-column pairs (ie each cell in the original table $Y$) with:

$$
H_k(i, j) = \frac{L_k(i) + C_k(j)}{\sqrt{2 \lambda_k \mu_k}}
$$

where $L_k(i)$ is the coordinate of row (site) $i$ on axis $k$ and $C_k(j)$ is the same for column (species) $j$ and $\mu_k = 1 + \sqrt{\lambda_k}$.

```{r}
mu <- 1 + sqrt(lambda)
```

The result is a matrix $H$ with $\omega = \sum_Y y_{ij} \neq 0$ rows (one row for each nonzero cell in $Y$, called correspondences) and $K$ columns (one column per principal axis of the CA):

```{r, fig.height=10, fig.width=4, echo=FALSE}
plotmat(r = sum(Y != 0), c = length(lambda),
        Yname = "italic(H)",
        rname = "italic(omega)",
        cname = "italic(K)",
        plot.margin.x = 10)
```

Let's compute $H$ from the CA scores with our example dataset.

```{r}
# Transform matrix to count table
Yfreq <- as.data.frame(as.table(Y))
colnames(Yfreq) <- c("row", "col", "Freq")

# Remove the cells with no observation
Yfreq0 <- Yfreq[-which(Yfreq$Freq == 0),]
Yfreq0$colind <- match(Yfreq0$col, colnames(Y)) # match index and species names

# Initialize results matrix
H <- matrix(nrow = nrow(Yfreq0), 
            ncol = length(lambda))

for (k in 1:length(lambda)) { # For each axis
  ind <- 1 # initialize row index
  for (obs in 1:nrow(Yfreq0)) { # For each observation
    i <- Yfreq0$row[obs]
    j <- Yfreq0$col[obs]
    H[ind, k] <- (L[i, k] + C[j, k])/sqrt(2*lambda[k]*mu[k])
    ind <- ind + 1
  }
}
```

```{r, include=FALSE}
# Compare results ----

# Get one observation from the coordinates computed above
i <- 1
j <- 3
jtxt <- colnames(Y)[j]
k <- 2

Htest <- H[Yfreq0$row == i & Yfreq0$col == jtxt, k]

# Compute one value with the formula
Hkij <- (L[i, k] + C[j, k])/sqrt(2*lambda[k]*mu[k])
Htest/Hkij

# Compare to scores computed with reciprocal.coa
H/rec_ca[, 1:(c-1)] # ade4
```


```{r}
# Test if the correspondences can be interpreted as a weighted mean
i <- 1
j <- 3
jtxt <- colnames(Y)[j]
k <- 2

# Get the H from reciprocal scaling
Htest <- H[Yfreq0$row == i & Yfreq0$col == jtxt, k]

# Use CA coordinates
Hkij <- (ca$li[i, k] + ca$co[j, k])/2

ca$co/sweep(ca$c1, 2, sqrt(lambda), "*")

# Compare
Htest/Hkij
```

On this plot, we plot the CA coordinates for site $i$ and for all species $j$ in site $i$ (scaled with a factor $\frac{\sqrt{\mu_k}}{\sqrt{2\lambda_k}}$. We surimpose the reciprocal scaling correspondences for site $i$ (purple dots).


```{r}
# Test if the correspondences can be interpreted as a weighted mean
i <- 1
k <- 1:2

# Get the H from reciprocal scaling
Hsub <- H[Yfreq0$row == i, k]
j <- Yfreq0$col[Yfreq0$row == i]
rownames(Hsub) <- paste0("s", i, "-", j)

# Get corresponding CA coordinates
# For site i
ca_i <- ca$li[i, k]*sqrt(mu[k])/sqrt(2*lambda[k])
# For species j presentin site i
ca_j <- sweep(ca$co[j, k], 2, sqrt(mu[k])/sqrt(2*lambda[k]), "*")

s.class(Hsub, 
        fac = factor(rep(i, nrow(Hsub))),
        wt = Yfreq0$Freq[Yfreq0$row == i],
        ppoints.col = "darkgrey")
s.label(Hsub, 
        plabels.optim = TRUE, add = TRUE,
        ppoints.col = "darkorchid",
        plabels.col = "darkorchid")
s.label(ca_i, 
        plabels.col = "cornflowerblue", add = TRUE)
s.label(ca_j, 
        plabels.col = "darkred", 
        ppoints.col = "darkred", 
        plabels.optim = TRUE,
        add = TRUE)
```

Below is a graphical demonstration that we can also compute ellipses from the CA scores, that look similar to the ellipses computed with reciprocal scaling: to do that, we compote the "CA-correspondences" with
$$\tilde{H}_k(i, j) = \frac{L1_k(i) + C_k(j)}{2}$$

```{r}
cali_rep <- c()
for (jiter in 1:length(j)) {
  cali_rep <- rbind(cali_rep, ca$li[i, ])
}

tst <- (ca$c1[j, ] + cali_rep)/2

s.class(ca$c1[j, ], 
        fac = factor(rep(i, nrow(ca$c1[j, ]))),
        wt = Yfreq0$Freq[Yfreq0$row == i],
        ppoints.col = "darkgrey")
s.label(ca$c1[j, ], 
        plabels.col = "darkred", 
        ppoints.col = "darkred", 
        plabels.optim = TRUE,
        add = TRUE)
s.class(tst, 
        fac = factor(rep(i, nrow(Hsub))),
        wt = Yfreq0$Freq[Yfreq0$row == i],
        ppoints.col = "darkorchid",
        add = TRUE)
s.label(ca$li[i, ], 
        plabels.col = "cornflowerblue", add = TRUE)

```

And if we were to compute the correspondences in the other biplot, it would be:
$$\tilde{H}_k(i, j) = \frac{L_k(i) + C1_k(j)}{2}$$

### From canonical correlation analysis

Canonical correlation analysis is the extension of correlation to multidimensional analysis and allows to find the coefficients to maximize the correlation between the columns of two matrices.

First, we compute the inflated tables $R$ ($\omega \times r$) and $C$ ($\omega \times c$) from $Y$ ($r \times c$).

We take the frequency table defined before and use it to compute the inflated tables (with weights):

```{r}
# Create indicator tables
tabR <- acm.disjonctif(as.data.frame(Yfreq0$row))
colnames(tabR) <- rownames(Y)
tabC <- acm.disjonctif(as.data.frame(Yfreq0$col))
colnames(tabC) <- colnames(Y)

# Get weights
wt <- Yfreq0$Freq
```

Below are the first lines of tables $R$ and $C$:

```{r, echo=FALSE}
knitr::kable(head(tabR, 30))
knitr::kable(head(tabC, 30))
```

Then, we perform a canonical correlation on the scaled tables $R_{scaled}$ and $C_{scaled}$. We find the coefficients $\rho$ and $\gamma$ maximizing the correlation between the scores $S_R = R_{scaled} \rho$ and $S_C = C_{scaled} \gamma$.

```{r}
# Center tables
tabR_scaled <- scalewt(tabR, wt, 
                       scale = FALSE)
tabC_scaled <- scalewt(tabC, wt, 
                       scale = FALSE)

res <- cancor(diag(sqrt(wt)) %*% tabR_scaled, 
              diag(sqrt(wt)) %*% tabC_scaled, 
              xcenter = FALSE, ycenter = FALSE)
# res gives the coefficients of the linear combinations that maximizes the correlation between the 2 dimensions
ncol(res$xcoef) # r-1 columns -> R_scaled is not of full rank
ncol(res$ycoef) # c-1 columns -> C_scaled is not of full rank

# Compute these scores from this coef
scoreR <- tabR_scaled[, 1:(r-1)]  %*% res$xcoef
scoreC <- tabC_scaled[, 1:(c-1)]  %*% res$ycoef
```

```{r}
dim(scoreR[, 1:20])
dim(ca$li)

scoreR[1:26, 1:20]/ca$li
```

We have $H = (S_R + S_C)_{scaled}$.

```{r}
# Get H
scoreRC <- scoreR[, 1:(c-1)] + scoreC[, 1:(c-1)] # here c-1 < r so c-1 axes
scoreRC_scaled <- scalewt(scoreRC, wt = wt) # normalisation à 1
```

```{r, include=FALSE}
# Check result
all(abs(scoreRC_scaled/H) - 1 < 10e-10)
```

### Plot

The scores $H$ are displayed below on the first two axes:

```{r, warning=FALSE, fig.height=6, fig.width=6}
#| code-fold: true
multiplot(indiv_row = H, 
          indiv_row_lab = paste0("site ", Yfreq0$row, "/", Yfreq0$col),
          row_color = "black", eig = lambda)
```

We can group those scores per species of per site:

```{r}
#| code-fold: true
# Plot with RC score = H
s.label(H,
        labels = paste(Yfreq0$row, Yfreq0$col, sep = "-"),
        plabels.optim = TRUE,
        plabels.col = "darkgrey",
        main = "Sites")
s.class(H, 
        wt = Yfreq0$Freq,
        fac = as.factor(Yfreq0$row),
        plabels.col = params$colsite,
        add = TRUE)

s.label(H,
        labels = paste(Yfreq0$row, Yfreq0$col, sep = "-"),
        plabels.optim = TRUE,
        plabels.col = "darkgrey",
        main = "Species")
s.class(H, 
        wt = Yfreq0$Freq,
        fac = as.factor(Yfreq0$colind),
        labels = colnames(Y),
        plabels.col = params$colspp,
        add = TRUE)
```
We will demonstrate more in-depth below how these groups are linked to the CA scores (means and variances can be computed directly from the CA scores.)


## Conditional means and variances of RC scores

::: {.callout-tip title="*TL;DR*"}
Once we have the correspondences scores, we can group them by row (site) or column (species) to compute conditional summary statistics:

-   conditional mean for site $i$ or species $j$ (for each axis $k$)
-   conditional variance for site $i$ or species $j$ (for each axis $k$)
-   conditional covariance for site $i$ or species $j$ (between axes $k$ and $l$)

These conditional statistics can be computed using $H_k(i,j)$ or using the CA scores:

**Formulas using** $H_k(i,j)$

The means are simply the weighted means of the $H_k(i, j)$ for a fixed $i$ or $j$ (corresponds to `meanfacwt`).
Similarly, the variances are weighted variances (can be computed with `meanvarwt`).


|                                  | Rows (sites)                                                                               | Columns (species)                                                                           |
|-------------------|------------------------------|------------------------|
| Mean for axis $k$                | $$m_k(i) = \frac{1}{y_{i\cdot}} \sum_{j = 1}^c y_{ij} H_k(i,j)$$                           | $$m_k(j) = \frac{1}{y_{\cdot j}} \sum_{i = 1}^r y_{ij} H_k(i,j)$$                           |
| Variance for axis $k$            | $$s_k^2(i) = \frac{1}{y_{i\cdot}} \sum_{j = 1}^c y_{ij} \left(H_k(i,j) - m_k(i)\right)^2$$ | $$s_k^2(j) = \frac{1}{y_{\cdot j}} \sum_{i = 1}^r y_{ij} \left(H_k(i,j) - m_k(j)\right)^2$$ |
| Covariance between axes $k$, $l$ | $$c_{kl}(i) = \frac{1}{y_{i \cdot}}\sum_{j=1}^c y_{ij}H_k(i, j)H_l(i, j) - m_k(i)m_l(i)$$  | $$c_{kl}(j) = \frac{1}{y_{\cdot j}}\sum_{i=1}^r y_{ij}H_k(i, j)H_l(i, j) - m_k(j)m_l(j)$$   |

**Formulas using CA scores**

Here, we can see that the means per species or site is equivalent to the CA coordinates (`co` and `li`) but with a proportionality a factor $\frac{\sqrt{\mu_k}}{\sqrt{2\lambda_k}}$.

The variances are equivalent to a "weighted variance" (variance developed formula: $\frac{\sum x_i^2}{n} - \bar{x}^2$). To explain better, we take the example of mean per rows (sites). The mean is $\sqrt{\lambda_k} L_k(i)$ ($L_k(i)$ = `li`). And there is a factor $\frac{1}{2\lambda_k \mu_k}$ compared to the variance of the species (`co` or $C_k(j)$) around this mean. For the mean per species, it is the same bit inverting the roles of `co` and `li`.

|                                  | Rows (sites)                                                                                                                                                                                | Columns (species)                                                                                                                                                                           |
|-------------------|------------------------------|------------------------|
| Mean for axis $k$                | $$m_k(i) = \frac{\sqrt{\mu_k}}{\sqrt{2\lambda_k}} L_k(i)$$                                                                                                                                  | $$m_k(j) = \frac{\sqrt{\mu_k}}{\sqrt{2\lambda_k}} C_k(j)$$                                                                                                                                  |
| Variance for axis $k$            | $$s^2_k(i) = \frac{1}{2\lambda_k\mu_k} \left( \frac{1}{y_{i \cdot}} \sum_{j = 1}^c \left(y_{ij} C_k^2(j) \right) - \lambda_k L_k^2(i) \right)$$                                             | $$s^2_k(j) = \frac{1}{2\lambda_k\mu_k} \left( \frac{1}{y_{\cdot j}} \sum_{i = 1}^r \left(y_{ij} L_k^2(i) \right) - \lambda_k C_k^2(j) \right)$$                                             |
| Covariance between axes $k$, $l$ | $$c_{kl}(i) = \frac{1}{2\sqrt{\lambda_k \lambda_l} \sqrt{\mu_k \mu_l}} \left( \frac{1}{y_{i \cdot}} \sum_{j = 1}^c y_{ij} C_k(j) C_l(j) - \sqrt{\lambda_k \lambda_l} L_k(i)L_l(i) \right)$$ | $$c_{kl}(j) = \frac{1}{2\sqrt{\lambda_k \lambda_l} \sqrt{\mu_k \mu_l}} \left( \frac{1}{y_{\cdot j}} \sum_{i = 1}^r y_{ij} L_k(i) L_l(i) - \sqrt{\lambda_k \lambda_l} C_k(j)C_l(j) \right)$$ |

***Note: formulas for the variances*** $s^2_k(i)$ and $s^2_k(j)$ are different from formula (15) in @thioulouse1992 (no square root at the denominator $2\lambda_k\mu_k$).
:::

We can compute the conditional means, variances and covariances using reciprocal scaling scores. There are 2 equivalent formulas for that:

-   using the $H_k(i,j)$ scores
-   using the CA coordinates

### Using $H_k(i,j)$

#### For rows = sites

We compute the weighted mean of $H_k(i,j)$ per row = site $i$.

$$
m_k(i) = \frac{1}{y_{i\cdot}} \sum_{j = 1}^c y_{ij} H_k(i,j)
$$

```{r}
# Get marginal counts
yi_ <- rowSums(Y)
y_j <- colSums(Y)
```

```{r}
# Initialize mean vector
mrows <- matrix(nrow = r, 
                ncol = c-1) 

for (i in 1:r) {
  # Get nonzero cells for site i
  rows <- which(Yfreq0$row == i)
  
  # Get scores for site i
  Hi <- H[rows, ]
  
  # Get counts for site i
  yij <- Yfreq0$Freq[rows]
  
  # Fill i-th row with site i mean score along each axis
  # (colSums sums all species j in site i)
  if (is.matrix(Hi)) { # There are several species in site i
    mrows[i, ] <- 1/yi_[i]*colSums(diag(yij) %*% Hi)
  } else { # Only one species in that site
    mrows[i, ] <- 1/yi_[i]*colSums(yij %*% Hi)
  }
  
}
```

```{r, include = FALSE}
# Check agreement with ade4 method
recscal_mean <- meanfacwt(rec_ca[, 1:(c-1)], 
                          rec_ca$Row, 
                          wt = rec_ca$Weight)

all(abs(recscal_mean/mrows) - 1 < 10e-10)
```

We compute the (weighted) variance of $H_k(i,j)$ per row = site $i$.

$$
s_k^2(i) = \frac{1}{y_{i\cdot}} \sum_{j = 1}^c y_{ij} \left(H_k(i,j) - m_k(i)\right)^2
$$ *NB: in the original article [@thioulouse1992] they use the compact variance formula* $s_k^2(i) = \frac{1}{y_{i\cdot}} \sum_{j = 1}^c \left(y_{ij} H_k(i,j)^2\right) - m_k(i)^2$ but I prefer the other formula.

```{r}
# Initialize mean vector
varrows <- matrix(nrow = r, 
                  ncol = c-1) 

for (i in 1:r) {
  # Get nonzero cells for site i
  rows <- which(Yfreq0$row == i)
  
  # Get scores for site i
  Hi <- H[rows, ]
  
  # Get counts for site i
  yij <- Yfreq0$Freq[rows]
  
  # Fill i-th row with site i variance along each axis
  varrows[i, ] <- 1/yi_[i]*colSums(diag(yij) %*% sweep(Hi, 2, mrows[i, ], "-")^2)
}
```

```{r, include = FALSE}
# Check agreement with ade4 method
li_rescsal <- s.class(rec_ca[,  1:2], 
                      wt = rec_ca$Weight, 
                      fac = rec_ca$Row,
                      plot = FALSE)
varrows_ax1 <- sapply(li_rescsal@stats$covvar, function(x) x[1,1])
varrows_ax1/varrows[, 1]
```

Finally, we can compute the covariance between scores on the several axes per row = site $i$. The covariance between two axes $k$ and $l$ of the multivariate space for site = row $i$ is defined as:

$$
c_{kl}(i) = \frac{1}{y_{i \cdot}}\sum_{j=1}^c y_{ij}H_k(i, j)H_l(i, j) - m_k(i)m_l(i)
$$

```{r}
k <- 1
l <- 2
```

Below, we compute the covariance between axes $k =$ `r k` and $l =$ `r l`.

```{r, include=FALSE}
# Compute covariance for site 3 only
i <- 3
rows <- which(Yfreq0$row == i)

covrow1 <- (1/yi_[i])*sum(Yfreq0$Freq[rows] * H[rows, k] * H[rows, l]) - mrows[i, k]*mrows[i, l]
```

```{r}
# Compute covariances for all sites

# Create a df containing intermediate computation and row = site identifier
df <- data.frame(var = Yfreq0$Freq*H[, k]*H[, l], 
                 by = Yfreq0$row)
# Get the sum over the different species for each site
sum_per_site <- aggregate(df$var, list(df$by), sum)$x

# Then vector-compute covariances
covrows <- (1/yi_)*sum_per_site - mrows[, k]*mrows[, l]
```

```{r, include=FALSE}
# Check agreement with ade4 method
li_rescsal <- s.class(rec_ca[,  1:2], 
                      wt = rec_ca$Weight, 
                      fac = rec_ca$Row,
                      plot = FALSE)
covrow_ade4 <- sapply(li_rescsal@stats$covvar, function(x) x[1,2])

# Check with simple computation method
covrow_ade4[i]/covrow1

# Check with vector method
covrow_ade4/covrows
```

#### For columns = species

Symetrically, we can compute the means, variances and covariances per columns = species $j$.

$$
m_k(j) = \frac{1}{y_{\cdot j}} \sum_{i = 1}^r y_{ij} H_k(i,j)
$$

$$
s_k^2(j) = \frac{1}{y_{\cdot j}} \sum_{i = 1}^r y_{ij} \left(H_k(i,j) - m_k(j)\right)^2
$$

$$c_{kl}(j) = \frac{1}{y_{\cdot j}}\sum_{i=1}^r y_{ij}H_k(i, j)H_l(i, j) - m_k(j)m_l(j)$$

```{r}
#| code-fold: true

# Compute means and variances ---
# Initialize mean vector
mcols <- matrix(nrow = c, 
                ncol = c-1) 
varcols <- matrix(nrow = c, 
                  ncol = c-1) 

for (j in 1:c) {
  # Get nonzero cells for species j
  cols <- which(Yfreq0$colind == j)
  
  # Get scores for species j
  Hj <- H[cols, ]
  
  # Get counts for species j
  yij <- Yfreq0$Freq[cols]
  
  # Fill j-th row with species j mean score/variance along each axis
  if (is.matrix(Hj)) { # There are several sites with this species
    mcols[j, ] <- 1/y_j[j]*colSums(diag(yij) %*% Hj)
    varcols[j, ] <- 1/y_j[j]*colSums(diag(yij) %*% sweep(Hj, 2, mcols[j, ], "-")^2)
  } else { # Only one site with that species
    mcols[j, ] <- 1/y_j[j]*colSums(yij %*% Hj)
    varcols[j, ] <- 1/y_j[j]*colSums(yij %*% (Hj - mcols[j, ])^2)
  }
  
}
```

```{r, include = FALSE}
# Compute covariance for species 3 only
j <- 4
cols <- which(Yfreq0$colind == j)

covcol1 <- (1/y_j[j])*sum(Yfreq0$Freq[cols] * H[cols, k] * H[cols, l]) - mcols[j, k]*mcols[j, l]
```

```{r}
#| code-fold: true

# Compute covariances ---

# Create a df containing intermediate computation and col = species identifier
df <- data.frame(var = Yfreq0$Freq*H[, k]*H[, l], 
                 by = Yfreq0$colind)
# Get the sum over the different sites for each species
sum_per_spp <- aggregate(df$var, list(df$by), sum)$x

# Then vector-compute covariances
covcols <- (1/y_j)*sum_per_spp - mcols[, k]*mcols[, l]
```

```{r, include = FALSE}
# Check agreement with ade4 method

# Mean
recscal_mean <- meanfacwt(rec_ca[, 1:(c-1)], 
                          rec_ca$Col, 
                          wt = rec_ca$Weight)

all(abs(recscal_mean/mcols) - 1 < 10e-10)

# Variance
col_rescsal <- s.class(rec_ca[,  1:2], 
                       wt = rec_ca$Weight, 
                       fac = rec_ca$Col)
varcols_ax1 <- sapply(col_rescsal@stats$covvar, function(x) x[1,1])
varcols_ax1/varcols[, 1]
# sp7 and 13 are Nan but it is because their variance is zero
varcols_ax1[7]
varcols[7, 1]

varcols_ax1[13]
varcols[13, 1]

# Both species are only present in one site
Y[, 7]
Y[, 13]
```

```{r, include=FALSE}
# Check agreement with ade4 method
co_rescsal <- s.class(rec_ca[,  1:2], 
                      wt = rec_ca$Weight, 
                      fac = rec_ca$Col,
                      plot = FALSE)
covcol_ade4 <- sapply(co_rescsal@stats$covvar, function(x) x[1,2])

# Check with simple computation method
covcol_ade4[j]/covcol1

# Check with vector method
covcol_ade4/covcols
# Species 7 and 13 are Nan because the variances are zero -> covariance too
```

### Using CA scores

#### For rows = sites

The conditional means and variances of rows scores on axis $k$ for row $i$ can also be computed as:

$$
m_k(i) = \frac{\sqrt{\mu_k}}{\sqrt{2\lambda_k}} L_k(i)
$$

$$
s^2_k(i) = \frac{1}{2\lambda_k\mu_k} \left( \frac{1}{y_{i \cdot}} \sum_{j = 1}^c \left(y_{ij} C_k^2(j) \right) - \lambda_k L_k^2(i) \right)
$$ ***Note: this formula is different from formula (15) in @thioulouse1992 (no square root at the denominator*** $2\lambda_k\mu_k$. Numerical computations suggest this is the correct formula.

```{r}
mrows2 <- sweep(ca$li, 2, sqrt(mu)/sqrt(2*lambda), "*")
```

```{r}
varrows2 <- matrix(nrow = r, 
                   ncol = c-1)

for (i in 1:r) {
  # Get CA scores for site i
  Li <- L[i, ]
  
  # Compute the part with the sum on Cj
  # we add all coordinates Cj^2 weighted by the number of observations on site i
  sumCj <- t(Y[i, ]) %*% as.matrix(C)^2

  # Fill i-th row with site i variance along each axis
  varrows2[i, ] <- 1/(2*lambda*mu)*((1/yi_[i])*sumCj - lambda*as.numeric(Li)^2)
}
```

```{r, include = FALSE}
mrows2/mrows
varrows2/varrows
```

We can also compute the covariance using CA coordinates:

$$
c_{kl}(i) = \frac{1}{2\sqrt{\lambda_k \lambda_l} \sqrt{\mu_k \mu_l}} \left( \frac{1}{y_{i \cdot}} \sum_{j = 1}^c y_{ij} C_k(j) C_l(j) - \sqrt{\lambda_k \lambda_l} L_k(i)L_l(i) \right)
$$

We will use the same values as above for $k$ and $l$: $k =$ `r k` and $l =$ `r l`.

```{r}
# Each site is multiplied by its score (Y*C[, k] by column) and then summed per site
sum_per_site <- rowSums(Y %*% diag(C[, k]) %*% diag(C[, l]))

covrows2 <- 1/(2*sqrt(lambda[k]*lambda[l])*sqrt(mu[k]*mu[l]))*(1/yi_*sum_per_site - sqrt(lambda[k]*lambda[l])*L[, k]*L[, l])
```

```{r, include = FALSE}
# Check result
covrows2/covrows
```

#### For columns = species

Same for the columns:

$$
m_k(j) = \frac{\sqrt{\mu_k}}{\sqrt{2\lambda_k}} C_k(j)
$$

$$
s^2_k(j) = \frac{1}{2\lambda_k\mu_k} \left( \frac{1}{y_{\cdot j}} \sum_{i = 1}^r \left(y_{ij} L_k^2(i) \right) - \lambda_k C_k^2(j) \right)
$$

$$
c_{kl}(j) = \frac{1}{2\sqrt{\lambda_k \lambda_l} \sqrt{\mu_k \mu_l}} \left( \frac{1}{y_{\cdot j}} \sum_{i = 1}^r y_{ij} L_k(i) L_l(i) - \sqrt{\lambda_k \lambda_l} C_k(j)C_l(j) \right)
$$

```{r}
mcols2 <- sweep(ca$co, 2, sqrt(mu)/sqrt(2*lambda), "*")
```

```{r}
varcols2 <- matrix(nrow = c, 
                   ncol = c-1)

for (j in 1:c) {
  # Get CA scores for species j
  Cj <- C[j, ]
  
  # Compute the part with the sum on Li
  sumLi <- t(Y[, j]) %*% as.matrix(L)^2

  # Fill i-th row with site i variance along each axis
  varcols2[j, ] <- 1/(2*lambda*mu)*((1/y_j[j])*sumLi - lambda*as.numeric(Cj)^2)
}
```

```{r}
# Each spp is multiplied by its score (Y*L[, k] by row) and then summed per spp
sum_per_spp <- colSums(diag(L[, k]) %*% diag(L[, l]) %*% Y)

covcols2 <- 1/(2*sqrt(lambda[k]*lambda[l])*sqrt(mu[k]*mu[l]))*(1/y_j*sum_per_spp - sqrt(lambda[k]*lambda[l])*C[, k]*C[, l])
```

```{r, include = FALSE}
mcols2/mcols
varcols2/varcols

# One inf column but it is because variance is zero
varcols[7, ]
varcols2[7, ]
```

```{r, include = FALSE}
# Check result
covcols2/covcols
```

## Conditional means and variances of R scores and C scores

::: {.callout-tip title="*TL;DR*"}

We can also compute compute means, variances and covariances with the R score or the C score.

We use the notation $m_k(i)$/$s^2_k(i)$ for the mean/variances of sites (index $i$) computed with the RC scores and $mR_k(i)$/$sR^2_k(i)$ for the mean/variances of sites computed with the R score. For the means computed for species, we use index $j$ and for the means/variances computed with C scores, we use $mC$/$sC^2$.

**Definition**

The definitions are simply the groupwise mean, variance and covariance.

+ using R score

|                       | Rows (sites)                      | Columns (species)        |
|-----------------------|-----------------------------------|--------------------------|
| Mean for axis $k$     | $$mR_k(i) = \frac{1}{y_{i\cdot}} \sum_j y_{ij} ~ \text{scoreR}_k(i, j)$$ | $$mR_k(j) = \frac{1}{y_{\cdot j}} \sum_i y_{ij} ~ \text{scoreR}_k(i, j)$$ |                                      
| Variance for axis $k$ |  $$sR^2_k(i) = \frac{1}{y_{\cdot j}} \sum_j y_{ij} ~ \text{scoreR}(i, j)^2 - mR_k(i)^2$$ | $$sR^2_k(j) = \frac{1}{y_{i \cdot}} \sum_j y_{ij} ~ \text{scoreR}(i, j)^2 - mR_k(j)^2$$ |
| Covariance between axes $k$, $l$ | | |

+ using C score

|                       | Rows (sites)                      | Columns (species)        |
|-----------------------|-----------------------------------|--------------------------|
| Mean for axis $k$     | $$mC_k(i) = \frac{1}{y_{i\cdot}} \sum_j y_{ij} ~ \text{scoreC}_k(i, j)$$ | $$mC_k(j) = \frac{1}{y_{\cdot j}} \sum_i y_{ij} ~ \text{scoreC}_k(i, j)$$ |
| Variance for axis $k$ | $$sC^2_k(i) = \frac{1}{y_{\cdot j}} \sum_j y_{ij} ~ \text{scoreC}(i, j)^2 - mC_k(i)^2$$ | $$sC^2_k(j) = \frac{1}{y_{i \cdot}} \sum_j y_{ij} ~ \text{scoreC}(i, j)^2 - mC_k(j)^2$$ |
| Covariance between axes $k$, $l$ | | |

**Formulas using CA score**

Now we can "prove" that the values defined above can be recomputed from the CA scores too.

+ using R score

|                       | Rows (sites)                        | Columns (species)        |
|-----------------------|-------------------------------------|--------------------------|
| Mean for axis $k$     | $\frac{1}{\sqrt{n}} L1_k(i)$ | $\frac{1}{\sqrt{n}} C_k(j)$ |
| Variance for axis $k$ | None                                | $$\frac{1}{n \lambda_k} \left( \frac{1}{y_{\cdot j}} \sum_{i = 1}^r \left(y_{ij} L_k^2(i) \right) - \lambda_k C_k^2(j) \right)$$ |
| Covariance between axes $k$, $l$ | None                     |                          |


+ using C score

|                       | Rows (sites)                      | Columns (species)        |
|-----------------------|-----------------------------------|--------------------------|
| Mean for axis $k$     | $L_k(i) \times \frac{1}{\sqrt{n}}$ | $C1_k(j) \times \frac{1}{\sqrt{n}}$ |
| Variance for axis $k$ |  $$\frac{1}{n \lambda_k} \left( \frac{1}{y_{i \cdot}} \sum_{j = 1}^c \left(y_{ij} C_k^2(j) \right) - \lambda_k L_k^2(i) \right)$$ | None                   |
| Covariance between axes $k$, $l$ |                         | None                    |



**Relation between RC scores and R score / C score**

Finally, we can relate the quantity computed above (with the definition or with the CA scores) with the means/variances/covariances computed with the RC scores.

+ using R score

|                       | Rows (sites)                      | Columns (species)        |
|-----------------------|-----------------------------------|--------------------------|
| Mean for axis $k$     | $$mR_k(i) = \frac{\sqrt{2}}{\sqrt{\mu_k n}} m_k(i)$$ | $$mR_k(j) = \frac{\sqrt{2 \lambda_k}}{\sqrt{\mu_k n}} m_k(j)$$ |
| Variance for axis $k$ | None | $$sR^2_k(j) = \frac{2 \mu_k}{n} s^2_k(j)$$ |
| Covariance between axes $k$, $l$ |  | |

+ using C score

|                       | Rows (sites)                      | Columns (species)        |
|-----------------------|-----------------------------------|--------------------------|
| Mean for axis $k$     | $$mC_k(i) =  \frac{\sqrt{2 \lambda_k}}{\sqrt{\mu_k n}} m_k(i)$$ | $$mC_k(j) = \frac{\sqrt{2}}{\sqrt{\mu_k n}} m_k(j) $$ |
| Variance for axis $k$ | $$sC^2_k(i) = \frac{2 \mu_k}{n} s^2_k(i)$$ | None |
| Covariance between axes $k$, $l$ |  | |

:::

This is not at all developed in the article of @thioulouse1992, but we could also use directly the R scores and the C scores to compute the groupwise (species- or sites-wise) mean and variances. In fact, the means and variances computed from the scoresR and scoresC should be related to the means and variances computed from $L_k(i)$ = `li` and $C_k(j)$ = `co`, since scoreR = `l1` and scoreC = `c1`.

We bother computing these because with CCA and dc-CA, we can't use directly RC scores. 

### For rows = sites (group per row)

```{r}
n <- sum(Y)
```

#### scoreC

Below is a graphical illustration of scoreC grouped by rows:

```{r}
s.class(scoreC,
        wt = Yfreq0$Freq,
        fac = as.factor(Yfreq0$row),
        ppoints.col = params$colspp,
        plabels.col = params$colsite, 
        main = "With C score (points = species scores)")
```

We compute the mean score per site from the **column** scores.

```{r}
mrowsC <- meanfacwt(scoreC, fac = Yfreq0$row, wt = Yfreq0$Freq)
```

We can relate this score to the CA scores: here, we have $mC_k(i) = \frac{L_k(i)}{\sqrt{n}}$

```{r}
mrowsC/(ca$li * 1/sqrt(n))
```

We can prove that equality: we begin by writing the weighted mean formula of scoreC:

$$mC_k(i) = \frac{1}{y_{i\cdot}} \sum_i y_{ij} ~ \text{scoreC}_k(i, j)$$
We replace scoreC with $\frac{C1_k(j)}{\sqrt{n}}$ (this equality remains to be proven):

$$mC_k(i) = \frac{1}{y_{i\cdot}} \sum_i y_{ij} \frac{C1_k(j)}{\sqrt{n}}$$

Then we remove $\frac{1}{n}$ from the sum:

$$mC_k(i) = \frac{1}{\sqrt{n}}\frac{1}{y_{i\cdot}} \sum_i y_{ij} C1_k(j)$$
By definition, the `li` are at the mean of the `c1` so we can replace $\frac{1}{y_{i\cdot}} \sum_i y_{ij} C1_k(j)$ with $L_k(i)$:

$$mC_k(i)  = \frac{1}{\sqrt{n}} L_k(i)$$

Now, we compare variances computed with the scoreC and variances computed with the scoreRC.

```{r}
varrowsC <- varfacwt(scoreC, fac = Yfreq0$row, wt = Yfreq0$Freq)

varrowsC %*% diag(1/(2*mu)) * n / varrows
```

Here too, we can prove the relationship above between the variance of the RC score and the variance of the C score.

$$
\begin{align}
sC_k^2(i) 
&= \frac{1}{y_{\cdot j}} \sum_j y_{ij} ~ \text{scoreC}(i, j)^2 - mC_k(i)^2\\ 
&= \frac{1}{y_{\cdot j}} \sum_j y_{ij} ~ \left(\frac{C1_k(j)}{\sqrt{n}}\right)^2 - \left(\frac{L_k(i)}{\sqrt{n}}\right)^2\\ 
&= \frac{1}{n} \left(\frac{1}{y_{\cdot j}} \sum_j y_{ij} ~ C1_k(j)^2 - L_k(i)^2\right)\\ 
&= \frac{1}{n} \left(\frac{1}{y_{\cdot j}} \sum_j y_{ij} ~ \left( \frac{C_k(j)}{\sqrt{\lambda_k}}\right)^2 - L_k(i)^2\right)\\ 
&= \frac{1}{n} \times \frac{1}{\lambda_k} \left(\frac{1}{y_{\cdot j}} \sum_j y_{ij} ~ C_k(j)^2 - \lambda_k L_k(i)^2\right)\\ 
&= \frac{1}{n} \times \frac{1}{\lambda_k} \times 2\lambda_k \mu_k \times \frac{1}{2\lambda_k \mu_k} \left(\frac{1}{y_{\cdot j}} \sum_j y_{ij} ~ C_k(j)^2 - \lambda_k L_k(i)^2\right)\\ 
sC_k^2(i) &= \frac{2\mu_k}{n} ~ s^2_k(i)
\end{align}
$$


#### scoreR

Below are R scores grouped by site:

```{r}
s.class(scoreR,
        wt = Yfreq0$Freq,
        fac = as.factor(Yfreq0$row),
        ppoints.col = params$colsite,
        plabels.col = params$colsite, 
        main = "With R score (points = sites scores)")
```

We compute the mean score per site from the **sites** scores (ie the unique scoreR, since all R values are superimposed for the same site).

```{r}
mrowsR <- meanfacwt(scoreR, fac = Yfreq0$row, wt = Yfreq0$Freq)
mrowsR * sqrt(n)/ca$l1

# NB: weighting doesn't make a difference because all weights are the same for the same row
mrowsR_nw <- meanfacwt(scoreR, fac = Yfreq0$row)
mrowsR_nw * sqrt(n)/ca$l1
```
We don't compute variances (they are null).

### For columns = species (group per column)

#### scoreC

Below are C scores grouped by column:

```{r}
s.class(scoreC,
        wt = Yfreq0$Freq,
        fac = as.factor(Yfreq0$colind),
        lab = colnames(Y),
        ppoints.col = params$colspp,
        plabels.col = params$colspp, 
        main = "With C score (points = species scores)")
```

All points are superimposed so we effectively select unique coordinates.

We compute the mean score per species from the **column** scores.

```{r}
mcolsC <- meanfacwt(scoreC, fac = Yfreq0$colind, wt = Yfreq0$Freq)

mcolsC * sqrt(n)/ca$c1
```

We don't compute variances (they are null).

#### scoreR

Below are R scores grouped by column:

```{r}
s.class(scoreR,
        wt = Yfreq0$Freq,
        fac = as.factor(Yfreq0$colind),
        lab = colnames(Y),
        ppoints.col = params$colsite,
        plabels.col = params$colspp, 
        main = "With R score (points = sites scores)")
```

We compute the mean score per species from the **row** scores.

```{r}
mcolsR <- meanfacwt(scoreR, fac = Yfreq0$colind, wt = Yfreq0$Freq)

mcolsR * sqrt(n)/ca$co
```

Now let's investigate the variances. We compute the variance from the R score:

```{r}
varcolsR <- varfacwt(scoreR, fac = Yfreq0$colind, wt = Yfreq0$Freq)

(varcolsR[, 1:20] %*% diag(1/(2*mu)) * n)/varcols
```

::: {.callout-note}
We can also compare means/variances extracted from the plot:

```{r}
#| code-fold: true
# Plot with RC score = H
xax <- 2
yax <- 3
s.label(scoreRC_scaled,
        xax = xax, yax = yax,
        labels = paste(Yfreq0$row, Yfreq0$col, sep = "-"),
        plabels.optim = TRUE,
        plabels.col = "darkgrey",
        main = "With RC scores (points = correspondences)")
sRC <- s.class(scoreRC_scaled, 
               xax = xax, yax = yax,
               wt = Yfreq0$Freq,
               fac = as.factor(Yfreq0$colind),
               labels = colnames(Y),
               plabels.col = params$colspp,
               add = TRUE)

# Plot with R score
sR <- s.class(scoreR,
              xax = xax, yax = yax,
              wt = Yfreq0$Freq,
              fac = as.factor(Yfreq0$colind),
              labels = colnames(Y),
              ppoints.col = params$colsite,
              plabels.col = params$colspp, 
              main = "With R score (points = sites scores)")
uniqueR <- as.data.frame(scoreR)
uniqueR$row <- Yfreq0$row
uniqueR <- uniqueR |> dplyr::distinct()
s.label(uniqueR[, c(xax, yax)]*sqrt(n),
        labels = uniqueR$row,
        plabels.optim = TRUE,
        plabels.col = params$colsite,
        ppoints.col = params$colsite,
        add = TRUE)
```


```{r}
fac <- sqrt(mu)/sqrt(2*lambda)

# Compare means
sRC$g2@stats$means/(sR@stats$means %*% diag(fac[c(xax, yax)])*sqrt(n)) # means are the same
# Means of scoreR per species = $co -> and means per H = fac*$co
```


```{r}
# Check that var measured on graph is the same as var computed with formula above
varcols[, xax]/sapply(sRC$g2@stats$covvar, function(x) x[1, 1])

# Compare var
(sapply(sRC$g2@stats$covvar, function(x) x[1, 1])*(2*mu[xax]/n))/(sapply(sR@stats$covvar, function(x) x[1, 1]))

(sapply(sRC$g2@stats$covvar, function(x) x[2, 2])*(2*mu[yax]/n))/(sapply(sR@stats$covvar, function(x) x[2, 2]))
```
:::


**Conclusion: using RC scores or using R scores only to compute species-wise means and variances is the same, but there are some multiplying factors.**


## Mean and variance plots

#### 1d plots

Let's plot these means and variances along the first axis for rows and columns:

```{r}
#| code-fold: true
# Define plotting function
plot_meanvar <- function(mean, var, groupnames, x = 1, col = "black")  {
  ggplot() +
  geom_point(aes(x = mean[, x], 
                 y = reorder(groupnames, -mean[, x])),
             colour = col, shape = 1) +
  geom_linerange(aes(x = mean[, x], 
                     y = reorder(groupnames, -mean[, x]),
                     xmin = mean[, x] - var[, x],
                     xmax = mean[, x] + var[, x]),
                 colour = col) +
  xlab(paste("Axis", x)) +
  theme_linedraw() +
  theme(axis.title.y = element_blank())
}
```

```{r, fig.width=9, fig.height=4}
#| code-fold: true
gr <- plot_meanvar(mrows, 3*sqrt(varrows), groupnames = rownames(Y),
                   col = params$colsite) +
  ggtitle("Rows (sites) mean and variance")
gc <- plot_meanvar(mcols, 3*sqrt(varcols), groupnames = colnames(Y),
                   col = params$colspp) +
  ggtitle("Columns (species) mean and variance")

grid.arrange(grobs = list(gr, gc), 
             nrow = 1)
```

```{r, include=FALSE}
# Same result with ade4 function
s1d.distri(score = ca$li[, 1], 
           dfdistri = as.data.frame(Y),
           ppoints.col = params$colspp,
           plabels.col = params$colspp)

s1d.distri(score = ca$co[, 1], 
           dfdistri = as.data.frame(t(Y)),
           ppoints.col = params$colsite,
           plabels.col = params$colsite)
```

```{r, fig.width=9, fig.height=4}
#| code-fold: true
gr <- ggplot() +
  geom_point(aes(x = covrows, y = reorder(rownames(Y), -covrows)),
             col = params$colsite) +
  xlab(paste("Covariances between axes", k, "and", l)) +
  theme_linedraw() +
  theme(axis.title.y = element_blank()) + 
  ggtitle("Rows (sites) covariances")

gc <- ggplot() +
  geom_point(aes(x = covcols, y = reorder(colnames(Y), -covcols)),
             col = params$colspp) +
  xlab(paste("Covariances between axes", k, "and", l)) +
  theme_linedraw() +
  theme(axis.title.y = element_blank()) + 
  ggtitle("Columns (species) covariances")

grid.arrange(grobs = list(gr, gc), nrow = 1)
```

#### 2d plots

Finally, we can plot these ellipses in 2 dimensions and summarize:

-   the conditional mean along axes $k$ and $l$
-   the conditional variance along axes $k$ and $l$
-   the conditional covariances between axes $k$ and $l$

For that, we need to plot the ellipse summarizing the Gaussian bivariate law for each group. For example, for site $i$ on axes $k$ and $l$ [reference blog post](https://www.xarg.org/2018/04/how-to-plot-a-covariance-error-ellipse/):

-   the mean vector is $\mu(i) = (m_k(i), m_l(i))$
-   the variance-covariance matrix is $\Sigma(i) = \begin{pmatrix} \sigma_k^2 ~ \sigma_{kl}\\ \sigma_{kl} ~ \sigma_{l}^2 \end{pmatrix}$

The simplest equation (coraviance zero, centered on the origin) is:

$$\left( \frac{x}{\sigma_k} \right)^2 + \left( \frac{y}{\sigma_l} \right)^2 = s$$ The semi-axes are of length $\sigma_k \sqrt{s}$ and $\sigma_l \sqrt{s}$. $s$ is the scaling factor allowing the ellipsis to encompass all data with a level of confidence $p$. We have $s = -2 \ln(1-p)$.

We can show that this equation is equivalent to

$$\begin{cases} x = \sigma_k \sqrt{s} \cos(t)\\ y = \sigma_l \sqrt{s} \sin(t)  \end{cases}$$ With $t$ being an angle between 0 and $2\pi$.

We write $a$ the length of the semi-major axis and $b$ the length of the semi-minor axis.

Now in case the values $H_k(i,j)$ and $H_l(i,j)$ are correlated, the covariance is not zero. Then, the semi-axes lengths are the square root of the eigenvaluesof $\Sigma(i)$. Geometrically, we have an angle $\theta$ between axis $k$ and the semi-major axis. $\theta$ can be computed as the direction of the first eigenvector $v_1$: $\theta = \arctan(v_1(2)/v_1(1))$.

The equation of the ellipsis is: $$\begin{cases} x = m_k(i) +  a  \cos(\theta) \cos(t) - b \sin(\theta) \sin(t) \\ y = m_l(i) + a \sin(\theta) \cos(t) + b \cos(\theta) \sin(t) \end{cases}$$ With $a = \sqrt{\lambda_1} \sqrt{s}$ and $b = \sqrt{\lambda_2} \sqrt{s}$. (Here we assumed the origin could be different from zero and shifted the center with $m_k(i)$ and $m_l(i)$.)

```{r}
#| code-fold: true
gaussian_ellipses <- function(vars, covars, means, x = 1, y = 2,
                              t = seq(0, 2*pi, 0.01), s = 1.5, ind = 1:nrow(vars)) {
  # Length of angles
  nt <- length(t)
  
  # Initialize matrix
  ellipses_mat <- matrix(nrow = length(t)*nrow(vars), 
                         ncol = 3)
  
  for (i in ind) {
    # Define variance covariance matrix
    covar <- covars[i]
    varx <- vars[i, x]
    vary <- vars[i, y]
    mat <- matrix(data = c(varx, covar,
                           covar, vary),
                  nrow = 2)
    
    # Eigenvalues
    vp <- eigen(mat)$values
    
    a <- s*sqrt(max(vp)) # semi major axis
    b <- s*sqrt(min(vp)) # semi minor axis
    
    # angle (inclination of ellipse from axe x)
    if(covar == 0 & varx >= vary){ # covariance is zero and varx is big
      theta <- 0
    }else if(covar == 0 & varx < vary){ # covariance is zero and varx is small
      theta <- pi/2
    }else{ # "normal" cases
      # get eigenvector associated to largest eigenvalue
      v1 <- eigen(mat)$vectors[,1]
      # theta is the angle of the eigenvector with x -> ie
      # arctan of the slope given by the first eigenvector
      theta <- atan(v1[2]/v1[1])
    }
    
    # x and y coordinates of ellipse
    xellipse <- means[i, x] + a*cos(theta)*cos(t) - b*sin(theta)*sin(t)
    yellipse <- means[i, y] + a*sin(theta)*cos(t) + b*cos(theta)*sin(t)
    
    indmin <- (i-1)*nt + 1
    indmax <- i*nt
    ellipses_mat[indmin:indmax, ] <- matrix(c(xellipse, yellipse, rep(i,nt)),
                                           ncol = 3)
  }
  res <- as.data.frame(ellipses_mat)
  colnames(res) <- c("xell", "yell", "group")
  return(res)
}
```

```{r}
#| code-fold: true

plot_ellipses <- function(var, covar, mean, groupnames, 
                          H, Yfreq0, col = "black", ellipses = 1:nrow(var), s = 1.5, x = 1, y = 2) {
  
  if (all(groupnames %in% Yfreq0$row)) {
    indivtype <- "row"
  } else {
    indivtype <- "colind"
  }
  
  rows <- which(Yfreq0[[indivtype]] %in% ellipses)
  
  # Get gaussian ellipses for the required groups
  ell <- gaussian_ellipses(var, covar, mean, s = s, x = k, y = l, ind = ellipses)
  
  # Plot ellipses from variance and covariance
  ggplot() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) +
    geom_point(aes(x = H[, x], y = H[, y], alpha = Yfreq0[[indivtype]] %in% ellipses),
               col = col,
               show.legend = FALSE) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.2)) +
    # Label ellipse points only
    geom_text_repel(aes(x = H[rows, x], y = H[rows, y], 
                        label = paste(Yfreq0$row[rows], Yfreq0$col[rows])),
                    col = col) +
    geom_polygon(data = ell, 
                 aes(x = xell, y = yell, group = group),
                 fill = col, col = col, alpha = 0.5) +
    theme_linedraw() +
    geom_label(aes(x = mean[, x], y = mean[, y], label = groupnames),
               colour = col) +
    xlab(paste("Axis", x)) +
    ylab(paste("Axis", y)) 
}
```

```{r, fig.width=6, fig.height=10}
#| code-fold: true

gr <- plot_ellipses(varrows, covrows, mrows, groupnames = rownames(Y), H = H, Yfreq0 = Yfreq0,
                    col = params$colsite) + ggtitle("Rows (sites)")

gc <- plot_ellipses(varcols, covcols, mcols, groupnames = colnames(Y), H = H, Yfreq0 = Yfreq0, 
                    col = params$colspp) + ggtitle("Columns (species)")

grid.arrange(grobs = list(gr, gc))
```

```{r, fig.width=6, fig.height=10}
#| code-fold: true

i <- c(2, 13, 18)
gr <- plot_ellipses(varrows, covrows, mrows, groupnames = rownames(Y), H = H, Yfreq0 = Yfreq0,
                    ellipses = i, col = params$colsite) + ggtitle("Rows (sites)")

j <- c(16, 12, 20) # sp21 is index 20
gc <- plot_ellipses(varcols, covcols, mcols, groupnames = colnames(Y), H = H, Yfreq0 = Yfreq0, 
                    ellipses = j, col = params$colspp) + ggtitle("Columns (species)")

grid.arrange(grobs = list(gr, gc))
```

```{r, include=FALSE}
# Check with ade4
s.class(rec_ca[,1:2],
        wt = rec_ca$Weight, 
        fac = rec_ca$Row,
        col = params$colsite)
s.class(rec_ca[,1:2], 
        wt = rec_ca$Weight, 
        fac = rec_ca$Col,
        col = params$colspp)
```

::: callout-note
Lets test formula (3) in @thioulouse1992 relating species and samples scores in CA.

The article says $L_k(i) = \sum_{j=1}^c p_{j|i} C_k(j)$ with $p_{j|i} = \frac{y_{ij}}{y_{i \cdot}}$.

```{r}
# Test formula for a given k
k <- 2

# compute pj|i
pjcondi <- sweep(Y, 1, yi_, "/")

# Compute the formula that should be equal th Lk(i)
res <- vector(mode="numeric", length=nrow(Y))
for (i in 1:nrow(Y)) {
  res[i] <- sum(pjcondi[i, ]*ca$co[, k])
}

res/ca$li[ ,k] # not equal to Lk(i)

res_lambda <- res*ca$eig[k]^(-1/2) # scale with 1/sqrt(lambda)

res_lambda/ca$li[ ,k] # Now it is equal to Lk(i)
```

We find that in fact the correct formula is

$$L_k(i) = \frac{1}{\sqrt{\lambda_k}}\sum_{j=1}^c p_{j|i} C_k(j)$$

With this equality we can indeed prove numerically the equivalence between mean computation using $H_k(i,j)$ (Equations (12) and (14) in @thioulouse1992) and using CA coordinates (computation in notebook not shown here).
:::
